---
title: "Context A/B: Penguin Visualization (No Context File)"
format:
  html:
    toc: true
    toc-depth: 2
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(palmerpenguins)
```

## Goal

Create a polished scatter plot exploring the relationship between bill length and body mass across penguin species.

## Data

```{r}
glimpse(penguins)
```

## Task

1. Remove rows with missing values.
2. Create a scatter plot of bill length (x) vs body mass (y), colored by species.
3. Style the plot for publication quality.
4. Write 2-3 sentences of descriptive interpretation.

## Starter Code

```{r}
#| eval: false

# TODO: clean the data

# TODO: build the scatter plot

# TODO: interpretation (as a comment or markdown text)
```

```{r}
# Clean the data
penguins_clean <- penguins %>%
  tidyr::drop_na(bill_length_mm, body_mass_g, species)

# Build the scatter plot
ggplot(
  penguins_clean,
  aes(x = bill_length_mm, y = body_mass_g, color = species)
) +
  geom_point(size = 2.6, alpha = 0.85) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "Bill length (mm)",
    y = "Body mass (g)",
    color = "Species",
    title = "Bill Length vs Body Mass in Palmer Penguins"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )
```

The scatter plot shows clear species-specific clustering in the bill length–body mass relationship. Gentoo penguins tend to have larger body mass and longer bills, while Adelie and Chinstrap cluster at lower body masses with shorter bills, indicating distinct morphological profiles by species.
