---
title: "AI Tools for Data Science and Statistics"
subtitle: "Project Presentations and Workflow Craft"
author: "Erik Westlund"
date: "March 12, 2026"
bibliography: references.bib
format:
  revealjs:
    slide-number: true
    transition: slide
    background-transition: fade
    self-contained: true
    css: slides.css
---

```{r}
#| include: false
if (requireNamespace("framework", quietly = TRUE)) {
  framework::scaffold()
} else {
  if (file.exists("../scaffold.R")) source("../scaffold.R")
}
```

## {background-color="#2c3e50"}

::: {style="color: white; font-size: 1.5em; text-align: center;"}
Session 4: Project Presentations and Workflow Craft
:::

## Plan for Today

1. Project presentations
2. Questions and discussion
3. Workflow tips for efficient, durable, low-friction work

## Project Presentations

Each group or student should briefly cover:

- What you are building or analyzing
- What worked
- What didn't
- Anything you found surprising, important, interesting, or instructive.

## Submitting Presentations

- You can email me your project at ewestlund@jhu.edu

## {background-color="#2c3e50"}

::: {style="color: white; font-size: 1.5em; text-align: center;"}
Workflow Craft
:::

## Don't Use A Forklift in the Gym

- You are a student, your goal is to learn
- If you want to get strong, don't use a forklift to lift weights and marvel over how much you can lift
- Everyone has access to these tools for cheap. Longterm, you will have a big comparative advantage if you can do everythign an LLM can do by yourself
- My hypothesis: the biggest "winners" in the AI era will be those who could live without LLMs, because they'll have the judgment to know when to use them and when not to, and the skills to do things themselves when they need to

## The Goal

Good workflow techniues:

- Make it easy to start/restart work
- Make it easy to inspect what changed
- Make it easy to recover from mistakes
- Make it easy to be safe with sensitive data
- Make it easy to leave a trail for your future self

## `.gitignore`

Two big reasons to care:

1. **Safety:** avoid accidental commits of data, secrets, logs, and generated files
2. **Hygiene:** reduce noise for both you and your tools

```text
data/
*.csv
.env
.Renviron
.claude/
.cursor/
```

## `.gitignore`: Practical Takeaway

- Keep sensitive or noisy files out of version control
- Keep your repo easier to search and review
- Give agents fewer irrelevant files to crawl

. . .

Helpful, but not security: `.gitignore` improves cleanliness, not access control.

## What Is `tmux`?

`tmux` is a terminal multiplexer: one persistent workspace for multiple terminal tasks.

- **Session:** a named project workspace
- **Window:** a tab inside that session
- **Pane:** a split inside a window

## Why `tmux` Helps

- Your work survives a closed laptop or dropped SSH connection
- You can keep editors, shells, REPLs, and agents open together
- You do not need to rebuild the same setup every day

This is mostly about persistence and fewer annoyances.

## A Simple `tmux` Pattern

Use a project launcher plus a few small aliases:

- `tm<project>`: start your standard project session
- `tma <session>`: attach to an existing session
- `tmk <session>`: kill a session you no longer need

Small ergonomics matter when you repeat them every day.

## Aliases Are Tiny Investments

- Short commands remove repeated friction
- Good aliases encode your preferred setup
- They help make local and remote work feel similar

## LLMs Help Automate Things That Used To Be Too "Expensive" to Automate

- LLMs can help you write scripts, aliases, and templates for common tasks
- When you have the thought "Wouldn't it be nice if..." ask yourself if an LLM can help you build that thing
- I do this all the time for workflow, down to having scripts that automate project setup.

## Keep A Paper Trail

- Keep a `docs/` or `notes/` directory for plans, prompts, and decision logs
- Save intermediate findings instead of trusting memory
- Let tools write summaries you can review later

Text files are cheap, searchable, and easy to revisit.

## Write Comments/Documentation You Otherwise Wouldn't

- We are all on tight deadlines
- Documentation is often overlooked or deferred
- Costs you later when you need to revisit
- Use tools to help generate documentation
- Make sure you give yourself time to review these for accuracy and bloat

## Write Tests You Otherwise Wouldn't

- Tests are a safety net for future changes
- They can be a sanity check for new tools or techniques
- They can be a way to capture assumptions and expected behavior
- Make sure you give yourself time to review these tests; LLMs will write useless stuff
- You can use LLMs to look for stupid tests, too.

## Checkpoints and Logs

- Commit after each meaningful step
- Use commits as rollback points when something goes sideways
- Keep short logs or notes so you can resume without rethinking everything

Good workflow reduces the cost of being interrupted.

## Sometimes You Should Just ****ing Do It Yourself

- I sometimes find myself working on a data visualization or similar task
- I know how to do it myself
- I spend 3x longer getting mad at an LLM for failing instead of just doing it myself

## Strong Pattern

- Design the project yourself
- Write out the function signatures and Roxygen blocks yourself
- Put comments in the functions saying what the function should do
- Let the LLM implement; this way, you are in charge

## Example

```{r}
#' Calculate the mean of a numeric vector
#' @param x A numeric vector
#' @return The mean of the vector
#' @export
calculate_mean <- function(x) {
  # Write a function to calculate the mean of a numeric vector
}
```

## Use LLMs to Organize Files

- I have moved away from going first to Zotero to organize literature reviews, notes, etc.
- Everything is a project with a directory
- LLMs can help organize things and summarize
- Not an excuse for reading, but having text of everything can really help you do things faster that would otherwise take forever.

## Use Alternative / Multiple Models To Look For Mistakes

- This is one of the lowest risk, high reward uses for LLMs
- Use a second model to review the work you produced
- Even use a third pass
- Don't trust these are always real mistakes; but even small catches can save you time and embarrassment later

## Use Models to Rubber Duck

- Rubber duck debugging is a classic technique where you explain your code to an inanimate object to find mistakes
- Explain your thinking to a model, even if you don't need it to solve a problem
- One of the biggest benefits of incorporating LLMs is I now spend a lot of time explaining my thoughts and goals, which brings clarity

## Models Aren't Therapists, But They Are Non-judgmental

- We all have gaps in our understanding, and we all make dumb mistakes
- It can be embarrassing to ask for help or admit confusion to colleagues
- LLMs can be a safe way to ask "stupid" questions or get feedback
- All caveats apply: they can be confidently wrong, and they won't know when they are wrong.

## Ask for Explanations

- Ask tools to explain what they are doing
- Catch misunderstandings before they compound
- Use one tool to review another when stakes are higher

Understanding the process is part of the workflow.

## Closing Thoughts

- AI Tools can accelerate work and learning, but they are not magic
- Good workflow practices can help you get the most out of these tools while avoiding common pitfalls
- Likewise, LLMs can help improve your workflow

. . .

- *But*: You  are responsible for your work, and you should understand it
