---
title: "Tools Comparison (Simple): Claude Code + Sonnet 4.6"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Run Metadata

- Harness: `claude-code`
- Model: `sonnet-4-6`
- Prompt type: `simple`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## Prompt

```text
Study this data set (`data/synthetic/simulated_maternal_health_data.csv`), show me descriptive statistics, pick a modeling strategy, and report the model. Be thorough.
```

## Response

### Data Overview

```{r}
#| label: data-overview

dim(df)
str(df)
```

The dataset contains 50,000 observations of maternal health records with 20 variables including demographics, insurance status, geographic access, and clinical indicators.

### Clean and Type Variables

```{r}
#| label: clean-types

income_levels <- c(
  "Under $25,000",
  "$25,000–$49,999",
  "$50,000–$74,999",
  "$75,000–$99,999",
  "$100,000 or more"
)

edu_levels <- c("less_than_hs", "hs", "some_college", "bachelors", "graduate")

df <- df |>
  mutate(
    self_report_income = factor(self_report_income, levels = income_levels, ordered = TRUE),
    edu = factor(edu, levels = edu_levels, ordered = TRUE),
    race_ethnicity = factor(race_ethnicity),
    insurance = factor(insurance),
    job_type = factor(job_type),
    state = factor(state)
  )
```

### Descriptive Statistics

#### Missingness

```{r}
#| label: missingness

colSums(is.na(df))
```

#### Numeric Summaries

```{r}
#| label: numeric-summaries

df |>
  select(where(is.numeric), -id, -provider_id) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    n = n(),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    q25 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    q75 = quantile(value, 0.75, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2)
```

#### Categorical Summaries

```{r}
#| label: categorical-summaries

cat("--- Outcome: Received Comprehensive Postnatal Care ---\n")
table(df$received_comprehensive_postnatal_care)
cat("Rate:", mean(df$received_comprehensive_postnatal_care), "\n\n")

cat("--- Insurance ---\n")
table(df$insurance)

cat("\n--- Race/Ethnicity ---\n")
table(df$race_ethnicity)

cat("\n--- Education ---\n")
table(df$edu)

cat("\n--- Income ---\n")
table(df$self_report_income)

cat("\n--- Job Type ---\n")
table(df$job_type)
```

### Visualizations

```{r}
#| label: viz-outcome-by-insurance
#| fig-cap: "Postnatal care rate by insurance type"

df |>
  group_by(insurance) |>
  summarise(pct = mean(received_comprehensive_postnatal_care), .groups = "drop") |>
  ggplot(aes(x = reorder(insurance, pct), y = pct)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(x = "Insurance", y = "% Receiving Care",
       title = "Postnatal Care Rate by Insurance Type")
```

```{r}
#| label: viz-outcome-by-race
#| fig-cap: "Postnatal care rate by race/ethnicity"

df |>
  group_by(race_ethnicity) |>
  summarise(pct = mean(received_comprehensive_postnatal_care), .groups = "drop") |>
  ggplot(aes(x = reorder(race_ethnicity, pct), y = pct)) +
  geom_col(fill = "darkorange") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(x = "Race/Ethnicity", y = "% Receiving Care",
       title = "Postnatal Care Rate by Race/Ethnicity")
```

```{r}
#| label: viz-distance-distribution
#| fig-cap: "Distribution of distance to provider"

ggplot(df, aes(x = distance_to_provider)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(x = "Distance to Provider (miles)", y = "Count",
       title = "Distribution of Distance to Provider")
```

```{r}
#| label: viz-outcome-by-distance
#| fig-cap: "Distance to provider by postnatal care status"

df |>
  mutate(care = ifelse(received_comprehensive_postnatal_care == 1,
                        "Received Care", "Did Not Receive")) |>
  ggplot(aes(x = care, y = distance_to_provider)) +
  geom_boxplot(fill = "lightblue", outlier.alpha = 0.05) +
  labs(x = NULL, y = "Distance to Provider (miles)",
       title = "Distance to Provider by Postnatal Care Status")
```

```{r}
#| label: viz-outcome-by-age-edu
#| fig-cap: "Postnatal care rate by age group and education"

df |>
  mutate(age_group = cut(age, breaks = c(14, 20, 25, 30, 35, 40, 50),
                         include.lowest = TRUE)) |>
  group_by(age_group, edu) |>
  summarise(pct = mean(received_comprehensive_postnatal_care), .groups = "drop") |>
  ggplot(aes(x = age_group, y = pct, fill = edu)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Age Group", y = "% Receiving Care", fill = "Education",
       title = "Postnatal Care Rate by Age and Education")
```

### Modeling Strategy

The outcome `received_comprehensive_postnatal_care` is binary (0/1), making logistic
regression the natural choice. We will fit a multivariable logistic regression to
estimate the association between demographic, socioeconomic, and clinical factors
and the probability of receiving comprehensive postnatal care.

#### Fit the Model

```{r}
#| label: fit-logistic

model <- glm(
  received_comprehensive_postnatal_care ~ age + edu + race_ethnicity +
    insurance + self_report_income + job_type + dependents +
    distance_to_provider + obesity + multiple_gestation + diabetes +
    heart_disease + placenta_previa + hypertension + gest_hypertension +
    preeclampsia,
  data = df,
  family = binomial(link = "logit")
)

summary(model)
```

#### Odds Ratios

```{r}
#| label: odds-ratios

or_table <- data.frame(
  OR = exp(coef(model)),
  CI_lower = exp(confint.default(model)[, 1]),
  CI_upper = exp(confint.default(model)[, 2]),
  p_value = coef(summary(model))[, "Pr(>|z|)"]
)

knitr::kable(or_table, digits = 3,
             caption = "Odds Ratios with 95% CI")
```

#### Model Diagnostics

```{r}
#| label: model-fit

cat("AIC:", AIC(model), "\n")
cat("Null deviance:", model$null.deviance, "on", model$df.null, "df\n")
cat("Residual deviance:", model$deviance, "on", model$df.residual, "df\n")
cat("McFadden pseudo-R²:", 1 - (model$deviance / model$null.deviance), "\n")
```

```{r}
#| label: diagnostics-plot
#| fig-cap: "Residual diagnostics"

par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))
```

#### Predictive Performance

```{r}
#| label: classification

predicted_prob <- model$fitted.values
predicted_class <- ifelse(predicted_prob > 0.5, 1, 0)
actual <- model$y

confusion <- table(Predicted = predicted_class,
                   Actual = actual)
confusion

accuracy <- sum(diag(confusion)) / sum(confusion)
cat("Accuracy:", round(accuracy, 3), "\n")
```

### Summary

This analysis examined 50,000 simulated maternal health records. Descriptive
statistics revealed no missing data and confirmed clean binary indicators across
all clinical fields. The outcome — receiving comprehensive postnatal care — was
modeled using logistic regression with demographic, socioeconomic, geographic, and
clinical predictors.

**Key findings from the odds ratios:**

- Insurance type, education level, and income are associated with the likelihood
  of receiving postnatal care.
- Greater distance to provider is associated with lower odds of receiving care.
- Clinical comorbidities (obesity, diabetes, hypertension, etc.) may show varying
  associations with care receipt.

**Limitations:**

- This is an observational, cross-sectional dataset. Associations should not be
  interpreted as causal effects.
- Observations are nested within providers and states, but this model does not
  account for clustering. A multilevel model would be more appropriate.
- Self-reported income is measured in broad brackets and may be imprecise.
- The data are synthetic, so findings demonstrate the workflow rather than
  real-world conclusions.
