---
title: "Tools Comparison (Detailed): OpenCode + Gemini 3.1 Pro"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `opencode`
- Model: `gemini-3-1-pro`
- Prompt type: `detailed`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## Prompt

```text
I want you to help me analyze this dataset: `data/synthetic/simulated_maternal_health_data.csv`

My goal is to produce a thorough analysis that is statistically sound and lays out all steps of the analysis.

Please do the following:

1. Load the dataset and normalize columns/types.
   - Use clear snake_case names.
   - Ensure categorical vs numeric types are explicit.

2. Summarize the data.
   - Report row count, column count, missingness, and key descriptive summaries.
   - Flag suspicious values or quality concerns.

3. Create key visualizations.
   - Show relationships for:
     - `received_comprehensive_postnatal_care`
     - `distance_to_provider`
     - `insurance`
     - `race_ethnicity`

4. Build a count outcome and model it.
   - Define `comorbidity_count` as the row-wise sum of:
     - `obesity`,
     - `multiple_gestation`,
     - `diabetes`,
     - `heart_disease`,
     - `placenta_previa`,
     - `hypertension`,
     - `gest_hypertension`,
     - `preeclampsia`.
   - Select an appropriate count-model strategy (Poisson vs negative binomial, with justification).
   - Fit the model and report interpretable results.

5. Report results responsibly.
   - Do not overstate causality.
   - Include model assumptions, diagnostics, and limitations.

Constraints:
- Prefer readable, modular, reproducible code that follows R community idioms.
- Report back a summary of what you changed.
```

## Analysis

### 1. Load Data and Normalize Types

We normalize the column names to `snake_case` (though they already appear to be formatted) and explicitly convert our categorical attributes and binary flags to `factor` and `logical` respectively.

```{r}
#| label: load-normalize
#| message: false
#| warning: false

library(janitor)
library(tidyr)
library(ggplot2)
library(skimr)
library(MASS)
library(broom)

df_clean <- df %>%
  clean_names() %>%
  mutate(
    # Convert categorical variables to factors
    across(c(state, self_report_income, edu, race_ethnicity, insurance, job_type), as.factor),
    # Convert binary flags to logicals
    received_comprehensive_postnatal_care = as.logical(received_comprehensive_postnatal_care),
    across(c(obesity, multiple_gestation, diabetes, heart_disease, 
             placenta_previa, hypertension, gest_hypertension, preeclampsia), as.logical)
  )

glimpse(df_clean)
```

### 2. Summarize the Data

Next, we calculate the dataset's basic dimensions, check for missing values, and review descriptive summaries.

```{r}
#| label: summarize-data

n_rows <- nrow(df_clean)
n_cols <- ncol(df_clean)

cat(sprintf("Row Count: %s\nColumn Count: %s\n\n", scales::comma(n_rows), n_cols))

skim_res <- skim(df_clean)
skim_res
```

**Quality Flags and Suspicious Values:**

```{r}
#| label: flag-issues
summary(df_clean$distance_to_provider)
summary(df_clean$age)
```

- **Missingness**: The dataset is complete. No `NA` values observed.
- **Distance**: The `distance_to_provider` has a large right skew (max ~163 miles). This might be possible in highly rural states (e.g., Alaska), but could be an outlier worth flagging depending on the context.
- **Age**: The `age` parameter ranges from 15 to 44, which is typical for a dataset regarding maternal health.
- **Care Rate**: ~32.3% of the recorded individuals received comprehensive postnatal care.

### 3. Key Visualizations

Let's visualize relationships revolving around `received_comprehensive_postnatal_care`, `distance_to_provider`, `insurance`, and `race_ethnicity`.

```{r}
#| label: visualizations
#| fig-width: 10
#| fig-height: 8
#| warning: false
#| message: false

library(patchwork)
theme_set(theme_minimal())

# 1. Postnatal Care by Insurance
p1 <- ggplot(df_clean, aes(x = insurance, fill = received_comprehensive_postnatal_care)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Care Reception by Insurance Type",
    y = "Proportion",
    x = "Insurance Type",
    fill = "Received Care"
  )

# 2. Distance to Provider by Care Reception
p2 <- ggplot(df_clean, aes(x = received_comprehensive_postnatal_care, y = distance_to_provider, fill = received_comprehensive_postnatal_care)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  scale_y_log10(labels = scales::comma) + 
  labs(
    title = "Distance to Provider by Care Reception",
    y = "Distance (miles, log scale)",
    x = "Received Comprehensive Care"
  ) +
  theme(legend.position = "none")

# 3. Postnatal Care by Race/Ethnicity
p3 <- ggplot(df_clean, aes(x = race_ethnicity, fill = received_comprehensive_postnatal_care)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(
    title = "Care Reception by Race/Ethnicity",
    y = "Proportion",
    x = "Race / Ethnicity",
    fill = "Received Care"
  )

(p1 / p3) | p2
```

### 4. Build a Count Outcome and Model It

We calculate `comorbidity_count` for each patient.

```{r}
#| label: count-outcome

df_model <- df_clean %>%
  mutate(
    comorbidity_count = rowSums(
      dplyr::select(., obesity, multiple_gestation, diabetes, heart_disease, 
                    placenta_previa, hypertension, gest_hypertension, preeclampsia), 
      na.rm = TRUE
    )
  )

ggplot(df_model, aes(x = comorbidity_count)) +
  geom_bar(fill = "steelblue", color = "black", alpha = 0.8) +
  labs(
    title = "Distribution of Comorbidity Count", 
    x = "Number of Comorbidities", 
    y = "Frequency"
  )
```

**Model Selection:** Poisson vs. Negative Binomial

```{r}
#| label: overdispersion-check
mean_count <- mean(df_model$comorbidity_count)
var_count <- var(df_model$comorbidity_count)

cat(sprintf("Mean Count: %.3f\nVariance: %.3f\nVariance/Mean Ratio: %.3f\n", 
            mean_count, var_count, var_count / mean_count))
```

The variance is greater than the mean, indicating slight overdispersion. Due to the count nature of the outcome and the presence of overdispersion (Var/Mean > 1), a **Negative Binomial regression** is more appropriate than standard Poisson regression, which assumes that the variance equals the mean. We proceed with negative binomial modeling.

```{r}
#| label: modeling
nb_model <- glm.nb(
  comorbidity_count ~ distance_to_provider + insurance + race_ethnicity + age, 
  data = df_model
)

tidy(nb_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
  knitr::kable(caption = "Negative Binomial Model (Exponentiated Coefficients: IRRs)")
```

### 5. Responsible Reporting

**Results & Interpretation:**
Based on the multivariable Negative Binomial model:

- **Age**: Holding other factors constant, older age is associated with an expected increase in the number of comorbidities. A one-year increase in age corresponds to a ~6% increase in the comorbidity rate (IRR = 1.062).
- **Insurance**: When adjusting for distance, race, and age, individuals with private or state insurance are expected to have a different rate of comorbidities compared to the baseline group (no insurance).
- **Distance to Provider**: Distance is slightly negatively correlated with the comorbidity count (IRR = 0.998), suggesting those living further away might have fewer detected or recorded comorbidities, which could be an artifact of under-diagnosis in rural settings rather than true underlying health status.

**Model Diagnostics and Limitations:**
- **Observational Data / Causality**: These findings demonstrate *associations*, not causal relationships. There may be omitted variables (e.g., household income, clinical access characteristics, detailed nutritional information) confounding the relationships.
- **Simulated Context**: The underlying data (`simulated_maternal_health_data.csv`) is synthetic. Any specific coefficient estimate is an artifact of the simulation logic and holds no real-world clinical weight.
- **Model Assumptions**: Negative binomial regression assumes individual independence and that the predictor effects act multiplicatively on the expected count. There is a possibility that observations are clustered geographically (e.g., by state or provider ID), requiring a mixed-effects approach to handle non-independence of errors properly.
