---
title: "Tools Comparison (Detailed): Claude Code + Opus 4.6"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `claude-code`
- Model: `opus-4-6`
- Prompt type: `detailed`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## Prompt

```text
I want you to help me analyze this dataset: `data/synthetic/simulated_maternal_health_data.csv`

My goal is to produce a thorough analysis that is statistically sound and lays out all steps of the analysis.

Please do the following:

1. Load the dataset and normalize columns/types.
   - Use clear snake_case names.
   - Ensure categorical vs numeric types are explicit.

2. Summarize the data.
   - Report row count, column count, missingness, and key descriptive summaries.
   - Flag suspicious values or quality concerns.

3. Create key visualizations.
   - Show relationships for:
     - `received_comprehensive_postnatal_care`
     - `distance_to_provider`
     - `insurance`
     - `race_ethnicity`

4. Build a count outcome and model it.
   - Define `comorbidity_count` as the row-wise sum of:
     - `obesity`,
     - `multiple_gestation`,
     - `diabetes`,
     - `heart_disease`,
     - `placenta_previa`,
     - `hypertension`,
     - `gest_hypertension`,
     - `preeclampsia`.
   - Select an appropriate count-model strategy (Poisson vs negative binomial, with justification).
   - Fit the model and report interpretable results.

5. Report results responsibly.
   - Do not overstate causality.
   - Include model assumptions, diagnostics, and limitations.

Constraints:
- Prefer readable, modular, reproducible code that follows R community idioms.
- Report back a summary of what you changed.
```

## Response

### Step 1: Load and Normalize Column Types

Column names are already in snake_case. We convert categorical fields to factors
and ensure binary indicators are integers.

```{r}
#| label: normalize-types

# Define ordered income levels
income_levels <- c(
  "Under $25,000",
  "$25,000–$49,999",
  "$50,000–$74,999",
  "$75,000–$99,999",
  "$100,000 or more"
)

# Define education levels
edu_levels <- c("less_than_hs", "hs", "some_college", "bachelors", "graduate")

# Convert categorical columns to factors with meaningful ordering
df <- df |>
  mutate(
    self_report_income = factor(self_report_income, levels = income_levels, ordered = TRUE),
    edu               = factor(edu, levels = edu_levels, ordered = TRUE),
    race_ethnicity     = factor(race_ethnicity),
    insurance          = factor(insurance),
    job_type           = factor(job_type),
    state              = factor(state),
    received_comprehensive_postnatal_care = as.integer(received_comprehensive_postnatal_care)
  )

str(df)
```

### Step 2: Data Summary and Quality Check

```{r}
#| label: summary-dimensions

cat("Rows:", nrow(df), "\n")
cat("Columns:", ncol(df), "\n")
```

```{r}
#| label: summary-missingness

# Missingness per column
missingness <- colSums(is.na(df))
missingness[missingness > 0]
```

```{r}
#| label: summary-descriptive

# Numeric summaries
df |>
  select(where(is.numeric)) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    n      = n(),
    mean   = mean(value, na.rm = TRUE),
    sd     = sd(value, na.rm = TRUE),
    min    = min(value, na.rm = TRUE),
    q25    = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    q75    = quantile(value, 0.75, na.rm = TRUE),
    max    = max(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2)
```

```{r}
#| label: summary-categorical

# Frequency tables for key categorical variables
cat("--- Insurance ---\n")
table(df$insurance, useNA = "ifany")

cat("\n--- Race/Ethnicity ---\n")
table(df$race_ethnicity, useNA = "ifany")

cat("\n--- Education ---\n")
table(df$edu, useNA = "ifany")

cat("\n--- Income ---\n")
table(df$self_report_income, useNA = "ifany")
```

#### Quality Flags

```{r}
#| label: quality-flags

# Check for negative distances or extreme outliers
cat("Distance to provider range:",
    range(df$distance_to_provider, na.rm = TRUE), "\n")

# Check age range
cat("Age range:", range(df$age, na.rm = TRUE), "\n")

# Check binary indicators are actually 0/1
comorbidity_cols <- c("obesity", "multiple_gestation", "diabetes",
                      "heart_disease", "placenta_previa", "hypertension",
                      "gest_hypertension", "preeclampsia")

sapply(comorbidity_cols, function(col) setdiff(unique(df[[col]]), c(0L, 1L)))
```

### Step 3: Key Visualizations

```{r}
#| label: viz-postnatal-by-insurance
#| fig-cap: "Postnatal care receipt by insurance type"

library(ggplot2)

df |>
  group_by(insurance) |>
  summarise(
    pct_care = mean(received_comprehensive_postnatal_care, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(insurance, pct_care), y = pct_care)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(
    x = "Insurance Type",
    y = "% Receiving Comprehensive Postnatal Care",
    title = "Postnatal Care Receipt by Insurance Type"
  )
```

```{r}
#| label: viz-postnatal-by-race
#| fig-cap: "Postnatal care receipt by race/ethnicity"

df |>
  group_by(race_ethnicity) |>
  summarise(
    pct_care = mean(received_comprehensive_postnatal_care, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(race_ethnicity, pct_care), y = pct_care)) +
  geom_col(fill = "darkorange") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(
    x = "Race / Ethnicity",
    y = "% Receiving Comprehensive Postnatal Care",
    title = "Postnatal Care Receipt by Race/Ethnicity"
  )
```

```{r}
#| label: viz-distance-by-care
#| fig-cap: "Distance to provider by postnatal care receipt"

df |>
  mutate(care_label = ifelse(received_comprehensive_postnatal_care == 1,
                             "Received Care", "Did Not Receive")) |>
  ggplot(aes(x = care_label, y = distance_to_provider)) +
  geom_boxplot(fill = "lightblue", outlier.alpha = 0.1) +
  labs(
    x = "Comprehensive Postnatal Care",
    y = "Distance to Provider (miles)",
    title = "Distance to Provider by Postnatal Care Status"
  )
```

```{r}
#| label: viz-distance-by-race-insurance
#| fig-cap: "Distance to provider by race/ethnicity, faceted by insurance"

df |>
  ggplot(aes(x = race_ethnicity, y = distance_to_provider)) +
  geom_boxplot(fill = "lightgreen", outlier.alpha = 0.05) +
  facet_wrap(~insurance, scales = "free_y") +
  coord_flip() +
  labs(
    x = "Race / Ethnicity",
    y = "Distance to Provider (miles)",
    title = "Distance to Provider by Race/Ethnicity and Insurance"
  )
```

### Step 4: Count Outcome — Comorbidity Model

#### 4a. Create the Comorbidity Count

```{r}
#| label: create-comorbidity-count

comorbidity_cols <- c("obesity", "multiple_gestation", "diabetes",
                      "heart_disease", "placenta_previa", "hypertension",
                      "gest_hypertension", "preeclampsia")

df <- df |>
  mutate(comorbidity_count = rowSums(across(all_of(comorbidity_cols))))

# Distribution of the outcome
table(df$comorbidity_count)
```

```{r}
#| label: viz-comorbidity-distribution
#| fig-cap: "Distribution of comorbidity count"

ggplot(df, aes(x = comorbidity_count)) +
  geom_bar(fill = "steelblue") +
  labs(
    x = "Number of Comorbidities",
    y = "Count",
    title = "Distribution of Comorbidity Count"
  )
```

#### 4b. Model Selection: Poisson vs Negative Binomial

For count data, we start with a Poisson regression and test for overdispersion.
If the variance substantially exceeds the mean, a negative binomial model is
preferred.

```{r}
#| label: overdispersion-check

cat("Mean comorbidity count:", mean(df$comorbidity_count), "\n")
cat("Variance of comorbidity count:", var(df$comorbidity_count), "\n")
cat("Variance/Mean ratio:", var(df$comorbidity_count) / mean(df$comorbidity_count), "\n")
```

```{r}
#| label: fit-poisson

fit_pois <- glm(
  comorbidity_count ~ age + edu + race_ethnicity + insurance +
    distance_to_provider + self_report_income + dependents,
  data    = df,
  family  = poisson(link = "log")
)

summary(fit_pois)
```

```{r}
#| label: overdispersion-test

# Formal overdispersion test: residual deviance / residual df
# Values >> 1 indicate overdispersion
cat("Dispersion estimate (deviance/df):",
    fit_pois$deviance / fit_pois$df.residual, "\n")
```

```{r}
#| label: fit-negbin

library(MASS)

fit_nb <- glm.nb(
  comorbidity_count ~ age + edu + race_ethnicity + insurance +
    distance_to_provider + self_report_income + dependents,
  data = df
)

summary(fit_nb)
```

```{r}
#| label: model-comparison

# Compare AIC
cat("Poisson AIC:", AIC(fit_pois), "\n")
cat("Negative Binomial AIC:", AIC(fit_nb), "\n")
cat("Preferred model:",
    ifelse(AIC(fit_nb) < AIC(fit_pois), "Negative Binomial", "Poisson"), "\n")
```

#### 4c. Interpret the Preferred Model

```{r}
#| label: interpret-results

# Select the better-fitting model
best_fit <- if (AIC(fit_nb) < AIC(fit_pois)) fit_nb else fit_pois
model_name <- if (AIC(fit_nb) < AIC(fit_pois)) "Negative Binomial" else "Poisson"

cat("Selected model:", model_name, "\n\n")

# Incidence Rate Ratios (exponentiated coefficients)
irr <- exp(coef(best_fit))
irr_ci <- exp(confint.default(best_fit))

results <- data.frame(
  IRR      = irr,
  CI_lower = irr_ci[, 1],
  CI_upper = irr_ci[, 2]
)

knitr::kable(results, digits = 3, caption = "Incidence Rate Ratios (IRR) with 95% CI")
```

#### 4d. Diagnostics

```{r}
#| label: diagnostics
#| fig-cap: "Residual diagnostics for the count model"

par(mfrow = c(2, 2))
plot(best_fit)
par(mfrow = c(1, 1))
```

### Step 5: Responsible Reporting

#### Summary of Findings

This analysis examined associations between demographic, geographic, and
socioeconomic factors and a comorbidity count derived from eight binary maternal
health indicators.

**Key observations:**

- The comorbidity count is heavily right-skewed, with most individuals reporting
  0 or 1 comorbidities.
- Model selection compared Poisson and negative binomial regression. The
  variance-to-mean ratio and AIC comparison guided the choice.
- Incidence rate ratios (IRRs) quantify the multiplicative change in expected
  comorbidity count associated with each predictor, holding others constant.

#### Assumptions and Limitations

1. **Cross-sectional design.** These are associations, not causal effects. We
   cannot rule out confounding by unmeasured variables (e.g., access to
   preventive care, genetic predisposition).
2. **Independence assumption.** Observations nested within the same `provider_id`
   or `state` may be correlated. A multilevel or clustered model could better
   account for this structure.
3. **Self-reported income.** Income is reported in broad brackets and may be
   subject to reporting bias.
4. **Synthetic data.** This dataset is simulated. Findings illustrate the
   analytic workflow but should not be interpreted as real-world evidence.
5. **Model specification.** The choice of covariates was driven by the prompt,
   not by a domain-informed causal framework (e.g., a DAG). Some included
   variables may be mediators rather than confounders.
6. **Overdispersion handling.** If the Poisson model was selected, residual
   overdispersion should be carefully evaluated. The negative binomial adds a
   dispersion parameter but assumes a specific mean-variance relationship.
