---
title: "Tools Comparison (Detailed): OpenCode + Opus 4.6"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `opencode`
- Model: `opus-4-6`
- Prompt type: `detailed`

## Setup

```{r}
#| label: setup
#| message: false

library(MASS)
library(dplyr)       # loaded after MASS so dplyr::select() is not masked
library(ggplot2)
library(tidyr)
library(broom)
library(scales)
library(patchwork)
library(knitr)
library(kableExtra)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## Prompt

```text
I want you to help me analyze this dataset: `data/synthetic/simulated_maternal_health_data.csv`

My goal is to produce a thorough analysis that is statistically sound and lays out all steps of the analysis.

Please do the following:

1. Load the dataset and normalize columns/types.
   - Use clear snake_case names.
   - Ensure categorical vs numeric types are explicit.

2. Summarize the data.
   - Report row count, column count, missingness, and key descriptive summaries.
   - Flag suspicious values or quality concerns.

3. Create key visualizations.
   - Show relationships for:
     - `received_comprehensive_postnatal_care`
     - `distance_to_provider`
     - `insurance`
     - `race_ethnicity`

4. Build a count outcome and model it.
   - Define `comorbidity_count` as the row-wise sum of:
     - `obesity`,
     - `multiple_gestation`,
     - `diabetes`,
     - `heart_disease`,
     - `placenta_previa`,
     - `hypertension`,
     - `gest_hypertension`,
     - `preeclampsia`.
   - Select an appropriate count-model strategy (Poisson vs negative binomial, with justification).
   - Fit the model and report interpretable results.

5. Report results responsibly.
   - Do not overstate causality.
   - Include model assumptions, diagnostics, and limitations.

Constraints:
- Prefer readable, modular, reproducible code that follows R community idioms.
- Report back a summary of what you changed.
```

## Analysis

### 1. Load and Normalize Data

Column names are already in `snake_case`. We convert categorical variables to explicit
factors with meaningful level ordering and recode the income variable to an ordered factor.

```{r}
#| label: normalize-data

# Define ordered income levels
income_levels <- c(
  "$0–$24,999",
  "$25,000–$49,999",
  "$50,000–$74,999",
  "$75,000–$99,999",
  "$100,000–$124,999",
  "$125,000–$149,999",
  "$150,000–$174,999",
  "$175,000+"
)

# Define ordered education levels
edu_levels <- c("less_than_hs", "hs", "some_college", "college", "post_grad")
edu_labels <- c("Less than HS", "High School", "Some College", "College", "Post-Grad")

df <- df |>
  mutate(
    # Binary outcome as factor
    received_comprehensive_postnatal_care = factor(
      received_comprehensive_postnatal_care,
      levels = c(0, 1),
      labels = c("No", "Yes")
    ),

    # Ordered factors
    self_report_income = factor(self_report_income, levels = income_levels, ordered = TRUE),
    edu = factor(edu, levels = edu_levels, labels = edu_labels, ordered = TRUE),

    # Unordered categorical
    race_ethnicity = factor(race_ethnicity),
    insurance = factor(insurance, levels = c("no_insurance", "state_provided", "private")),
    job_type = factor(job_type),
    state = factor(state),
    provider_id = factor(provider_id),

    # Binary indicators remain numeric (0/1) for modeling
    across(
      c(obesity, multiple_gestation, diabetes, heart_disease,
        placenta_previa, hypertension, gest_hypertension, preeclampsia),
      as.integer
    )
  )
```

### 2. Data Summary

#### Dimensions and Missingness

```{r}
#| label: dimensions

tibble(
  Metric = c("Rows", "Columns", "Total cells", "Total missing values"),
  Value = c(
    format(nrow(df), big.mark = ","),
    ncol(df),
    format(nrow(df) * ncol(df), big.mark = ","),
    sum(is.na(df))
  )
) |>
  kable(align = "lr") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r}
#| label: missingness

missingness <- df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  mutate(pct_missing = n_missing / nrow(df) * 100) |>
  filter(n_missing > 0)

if (nrow(missingness) == 0) {
  cat("No missing values detected in any column.\n")
} else {
  missingness |>
    arrange(desc(n_missing)) |>
    kable(digits = 2) |>
    kable_styling(bootstrap_options = "striped", full_width = FALSE)
}
```

#### Descriptive Statistics — Numeric Variables

```{r}
#| label: numeric-summary

df |>
  select(age, dependents, distance_to_provider) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    n = n(),
    mean = mean(value),
    sd = sd(value),
    min = min(value),
    q25 = quantile(value, 0.25),
    median = median(value),
    q75 = quantile(value, 0.75),
    max = max(value),
    .groups = "drop"
  ) |>
  kable(digits = 2) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Descriptive Statistics — Categorical Variables

```{r}
#| label: categorical-summary

cat_vars <- c(
  "received_comprehensive_postnatal_care", "self_report_income",
  "edu", "race_ethnicity", "insurance", "job_type"
)

cat_summary <- df |>
  select(all_of(cat_vars)) |>
  mutate(across(everything(), as.character)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "level") |>
  count(variable, level) |>
  group_by(variable) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()

cat_summary |>
  kable(digits = 1, col.names = c("Variable", "Level", "N", "%")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top")
```

#### Comorbidity Prevalence

```{r}
#| label: comorbidity-prevalence

comorbidity_vars <- c(
  "obesity", "multiple_gestation", "diabetes", "heart_disease",
  "placenta_previa", "hypertension", "gest_hypertension", "preeclampsia"
)

df |>
  select(all_of(comorbidity_vars)) |>
  pivot_longer(everything(), names_to = "condition", values_to = "present") |>
  group_by(condition) |>
  summarise(
    n_present = sum(present),
    prevalence_pct = mean(present) * 100,
    .groups = "drop"
  ) |>
  arrange(desc(prevalence_pct)) |>
  kable(digits = 1, col.names = c("Condition", "N Present", "Prevalence (%)")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Data Quality Notes

```{r}
#| label: quality-flags

# Check age range
cat("Age range:", range(df$age), "\n")

# Check distance for extreme values
dist_q99 <- quantile(df$distance_to_provider, 0.99)
n_extreme_dist <- sum(df$distance_to_provider > 100)
cat("Distance 99th percentile:", round(dist_q99, 1), "miles\n")
cat("Observations with distance > 100 miles:", n_extreme_dist,
    paste0("(", round(n_extreme_dist / nrow(df) * 100, 2), "%)"), "\n")

# Check dependents range
cat("Dependents range:", range(df$dependents), "\n")
```

> **Quality assessment:** No missing data. Age range (15–44) is plausible for a
> maternal health dataset. Distance to provider is right-skewed with a small
> number of extreme values (> 100 miles), which is realistic for rural settings
> but should be monitored during modeling. All binary comorbidity indicators are
> coded 0/1 with no out-of-range values.

### 3. Key Visualizations

#### Postnatal Care by Insurance Status

```{r}
#| label: fig-care-insurance
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Receipt of comprehensive postnatal care by insurance status"

df |>
  count(insurance, received_comprehensive_postnatal_care) |>
  group_by(insurance) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = insurance, y = pct, fill = received_comprehensive_postnatal_care)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2", name = "Postnatal Care") +
  labs(
    x = "Insurance Status",
    y = "Proportion",
    title = "Postnatal Care Receipt by Insurance Status"
  ) +
  theme_minimal(base_size = 13)
```

#### Postnatal Care by Race/Ethnicity

```{r}
#| label: fig-care-race
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Receipt of comprehensive postnatal care by race/ethnicity"

df |>
  count(race_ethnicity, received_comprehensive_postnatal_care) |>
  group_by(race_ethnicity) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = reorder(race_ethnicity, -pct), y = pct,
             fill = received_comprehensive_postnatal_care)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2", name = "Postnatal Care") +
  labs(
    x = "Race / Ethnicity",
    y = "Proportion",
    title = "Postnatal Care Receipt by Race/Ethnicity"
  ) +
  theme_minimal(base_size = 13)
```

#### Distance to Provider Distribution

```{r}
#| label: fig-distance
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Distribution of distance to provider, faceted by postnatal care receipt"

ggplot(df, aes(x = distance_to_provider, fill = received_comprehensive_postnatal_care)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  scale_x_continuous(trans = "log1p", breaks = c(0, 1, 5, 10, 25, 50, 100)) +
  scale_fill_brewer(palette = "Set2", name = "Postnatal Care") +
  facet_wrap(~ received_comprehensive_postnatal_care, ncol = 1, scales = "free_y") +
  labs(
    x = "Distance to Provider (miles, log scale)",
    y = "Count",
    title = "Distance to Provider by Postnatal Care Status"
  ) +
  theme_minimal(base_size = 13)
```

#### Distance to Provider by Insurance and Race

```{r}
#| label: fig-distance-insurance-race
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Distance to provider by insurance status and race/ethnicity"

p1 <- ggplot(df, aes(x = insurance, y = distance_to_provider)) +
  geom_boxplot(fill = "#a6cee3", outlier.alpha = 0.1) +
  scale_y_log10() +
  labs(x = "Insurance", y = "Distance (log scale)", title = "By Insurance") +
  theme_minimal(base_size = 12)

p2 <- ggplot(df, aes(x = race_ethnicity, y = distance_to_provider)) +
  geom_boxplot(fill = "#b2df8a", outlier.alpha = 0.1) +
  scale_y_log10() +
  labs(x = "Race / Ethnicity", y = NULL, title = "By Race/Ethnicity") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p1 + p2
```

### 4. Count Model: Comorbidity Count

#### Construct the Outcome

```{r}
#| label: comorbidity-count

df <- df |>
  mutate(
    comorbidity_count = obesity + multiple_gestation + diabetes +
      heart_disease + placenta_previa + hypertension +
      gest_hypertension + preeclampsia
  )

cat("Comorbidity count distribution:\n")
table(df$comorbidity_count) |> print()

cat("\nMean:", round(mean(df$comorbidity_count), 3),
    " Variance:", round(var(df$comorbidity_count), 3), "\n")
```

#### Model Selection: Poisson vs. Negative Binomial

For count data, Poisson regression assumes the mean equals the variance
(equidispersion). If the variance exceeds the mean, overdispersion is present
and a negative binomial model is more appropriate because it adds a dispersion
parameter.

```{r}
#| label: dispersion-check

mean_cc <- mean(df$comorbidity_count)
var_cc <- var(df$comorbidity_count)
cat("Mean:", round(mean_cc, 3), "\n")
cat("Variance:", round(var_cc, 3), "\n")
cat("Variance / Mean ratio:", round(var_cc / mean_cc, 3), "\n")
```

```{r}
#| label: fit-poisson

fit_pois <- glm(
  comorbidity_count ~ age + self_report_income + edu + race_ethnicity +
    insurance + job_type + dependents + distance_to_provider +
    received_comprehensive_postnatal_care,
  data = df,
  family = poisson(link = "log")
)
```

```{r}
#| label: overdispersion-test

# Residual deviance / df as quick overdispersion check
deviance_ratio <- deviance(fit_pois) / df.residual(fit_pois)
cat("Residual deviance / df:", round(deviance_ratio, 3), "\n")
cat("If substantially > 1, overdispersion is present.\n")
```

```{r}
#| label: fit-negbin

fit_nb <- glm.nb(
  comorbidity_count ~ age + self_report_income + edu + race_ethnicity +
    insurance + job_type + dependents + distance_to_provider +
    received_comprehensive_postnatal_care,
  data = df
)

cat("Negative binomial theta:", round(fit_nb$theta, 3), "\n")
cat("(Smaller theta = more overdispersion; as theta -> Inf, NB -> Poisson)\n")
```

#### Model Comparison

```{r}
#| label: model-comparison

aic_comp <- tibble(
  Model = c("Poisson", "Negative Binomial"),
  AIC = c(AIC(fit_pois), AIC(fit_nb)),
  `Log-Likelihood` = c(logLik(fit_pois), logLik(fit_nb))
)

aic_comp |>
  kable(digits = 1) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r}
#| label: lr-test

# Likelihood ratio test: NB vs Poisson (NB nests Poisson when theta -> Inf)
lr_stat <- 2 * (logLik(fit_nb) - logLik(fit_pois))
lr_pval <- pchisq(as.numeric(lr_stat), df = 1, lower.tail = FALSE)
cat("LR test statistic:", round(as.numeric(lr_stat), 1), "\n")
cat("p-value:", format.pval(lr_pval), "\n")
```

> **Decision:** We select the negative binomial model based on (1) the
> variance-to-mean ratio exceeding 1, (2) lower AIC, and (3) a significant
> likelihood ratio test rejecting the Poisson constraint.

#### Negative Binomial Results

```{r}
#| label: nb-summary

summary(fit_nb)
```

#### Interpretable Results: Incidence Rate Ratios

```{r}
#| label: irr-table

irr_table <- tidy(fit_nb, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    )
  ) |>
  select(term, IRR = estimate, `2.5%` = conf.low, `97.5%` = conf.high,
         p.value, significance)

irr_table |>
  kable(digits = 3) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Coefficient Plot

```{r}
#| label: fig-coef-plot
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Incidence rate ratios from negative binomial model (exponentiated coefficients with 95% CI)"

irr_plot_data <- tidy(fit_nb, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(term = reorder(term, estimate))

ggplot(irr_plot_data, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(
    x = "Incidence Rate Ratio (95% CI)",
    y = NULL,
    title = "Negative Binomial Model: Incidence Rate Ratios"
  ) +
  theme_minimal(base_size = 13)
```

### 5. Diagnostics and Responsible Reporting

#### Model Diagnostics

```{r}
#| label: fig-diagnostics
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Residual diagnostics for the negative binomial model"

diag_df <- tibble(
  fitted = fitted(fit_nb),
  pearson_resid = residuals(fit_nb, type = "pearson"),
  deviance_resid = residuals(fit_nb, type = "deviance")
)

p_resid_fitted <- ggplot(diag_df, aes(x = fitted, y = pearson_resid)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "blue", linewidth = 0.8) +
  labs(x = "Fitted Values", y = "Pearson Residuals",
       title = "Residuals vs Fitted") +
  theme_minimal(base_size = 12)

p_resid_hist <- ggplot(diag_df, aes(x = deviance_resid)) +
  geom_histogram(bins = 60, fill = "steelblue", alpha = 0.7) +
  labs(x = "Deviance Residuals", y = "Count",
       title = "Deviance Residual Distribution") +
  theme_minimal(base_size = 12)

p_resid_fitted + p_resid_hist
```

```{r}
#| label: rootogram
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Observed vs expected counts under the negative binomial model"

# Compare observed vs predicted count frequencies
obs_freq <- table(df$comorbidity_count)
pred_probs <- sapply(as.integer(names(obs_freq)), function(k) {
  mean(dnbinom(k, mu = fitted(fit_nb), size = fit_nb$theta))
})

rootogram_df <- tibble(
  count = as.integer(names(obs_freq)),
  observed = as.numeric(obs_freq) / nrow(df),
  expected = pred_probs
)

rootogram_df |>
  pivot_longer(cols = c(observed, expected), names_to = "type", values_to = "proportion") |>
  ggplot(aes(x = count, y = proportion, fill = type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("observed" = "steelblue", "expected" = "coral"),
                    name = NULL) +
  labs(
    x = "Comorbidity Count",
    y = "Proportion",
    title = "Observed vs. Expected Frequencies (Negative Binomial)"
  ) +
  theme_minimal(base_size = 13)
```

#### Assumptions and Limitations

**Model assumptions:**

1. **Conditional independence:** Observations are assumed independent given
   covariates. However, patients are nested within providers (`provider_id`)
   and states, introducing potential clustering. A mixed-effects model (e.g.,
   `glmer.nb()`) could account for this, but was not pursued here for
   parsimony.

2. **Correct mean structure:** We assume the log-linear specification captures
   the true relationship between covariates and the comorbidity count. No
   interactions or nonlinear terms for continuous predictors (e.g., age,
   distance) were included. Residual diagnostics above provide a visual check.

3. **Negative binomial distributional assumption:** The NB model adds a
   dispersion parameter relative to Poisson. The observed-vs-expected count
   comparison above assesses goodness-of-fit.

**Limitations:**

- **Cross-sectional, observational data.** All associations should be
  interpreted as correlations, not causal effects. Unmeasured confounders
  (e.g., access to prenatal care, health literacy, genetic factors) are likely
  present.

- **Self-reported income.** Subject to reporting bias and may not reflect
  actual economic resources (wealth, debt, cost of living).

- **Synthetic data.** This dataset is simulated. Results characterize the
  data-generating process, not real-world maternal health patterns.

- **No adjustment for multiple comparisons.** With many predictors, some
  nominally significant results may be spurious.

- **Clustering ignored.** Patients within the same provider or state may share
  unobserved characteristics. Standard errors may be underestimated.

---

## Summary of Changes

1. **Loaded and normalized** the 50,000-row dataset. Converted categorical
   columns (`self_report_income`, `edu`, `race_ethnicity`, `insurance`,
   `job_type`, `state`, `provider_id`) to factors with explicit levels.
   Ordered factors were used for income and education. The binary outcome
   `received_comprehensive_postnatal_care` was converted to a labeled factor.

2. **Summarized** the data: 50,000 rows, 20 columns, zero missing values.
   Reported descriptive statistics for all numeric and categorical variables.
   Flagged the right-skewed distance distribution and confirmed all binary
   indicators were within expected ranges.

3. **Created visualizations** showing postnatal care receipt by insurance
   status and race/ethnicity, distance-to-provider distributions by care
   status, and boxplots of distance by insurance and race.

4. **Built `comorbidity_count`** as the sum of 8 binary indicators. Compared
   Poisson and negative binomial models using AIC and a likelihood ratio test.
   The negative binomial was selected due to overdispersion. Reported
   incidence rate ratios with 95% confidence intervals and a coefficient plot.

5. **Reported responsibly** with residual diagnostics, an observed-vs-expected
   count comparison, and explicit discussion of assumptions (independence,
   log-linear mean, NB distribution) and limitations (observational design,
   self-report bias, synthetic data, clustering).
