---
title: "Tools Comparison (Simple): OpenCode + Codex GPT-5.3"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

if (requireNamespace("framework", quietly = TRUE)) {
  framework::scaffold()
} else {
  if (file.exists("../../../scaffold.R")) source("../../../scaffold.R")
}
```

## Run Metadata

- Harness: `opencode`
- Model: `codex-gpt-5-3`
- Prompt type: `simple`

## Setup

```{r}
#| label: setup
#| message: false

library(splines)

df <- data_read("data/synthetic/simulated_maternal_health_data.csv")

analysis_df <- df |>
  mutate(
    received_comprehensive_postnatal_care = as.integer(received_comprehensive_postnatal_care),
    state = factor(state),
    self_report_income = factor(self_report_income),
    edu = factor(edu),
    race_ethnicity = factor(race_ethnicity),
    insurance = factor(insurance),
    job_type = factor(job_type)
  )

binary_vars <- c(
  "obesity", "multiple_gestation", "diabetes", "heart_disease",
  "placenta_previa", "hypertension", "gest_hypertension", "preeclampsia"
)

auc_fast <- function(y, p) {
  r <- rank(p)
  n_pos <- sum(y == 1)
  n_neg <- sum(y == 0)
  (sum(r[y == 1]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

classification_metrics <- function(y_true, y_hat) {
  tp <- sum(y_true == 1 & y_hat == 1)
  tn <- sum(y_true == 0 & y_hat == 0)
  fp <- sum(y_true == 0 & y_hat == 1)
  fn <- sum(y_true == 1 & y_hat == 0)

  tibble(
    accuracy = (tp + tn) / length(y_true),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp),
    precision = tp / (tp + fp)
  )
}
```

## Prompt

```text
Study this data set (`data/synthetic/simulated_maternal_health_data.csv`), show me descriptive statistics, pick a modeling strategy, and report the model. Be thorough.
```

## Data Overview

```{r}
#| label: overview

tibble(
  n_rows = nrow(analysis_df),
  n_columns = ncol(analysis_df),
  n_states = n_distinct(analysis_df$state),
  n_providers = n_distinct(analysis_df$provider_id)
)
```

```{r}
#| label: missingness

analysis_df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_n") |>
  mutate(missing_pct = 100 * missing_n / nrow(analysis_df)) |>
  arrange(desc(missing_n))
```

## Descriptive Statistics

### Outcome Distribution

```{r}
#| label: outcome-distribution

analysis_df |>
  summarise(
    prevalence = mean(received_comprehensive_postnatal_care),
    n_positive = sum(received_comprehensive_postnatal_care),
    n_negative = sum(1 - received_comprehensive_postnatal_care)
  )
```

### Continuous Variables

```{r}
#| label: continuous-summary

analysis_df |>
  summarise(
    age_mean = mean(age),
    age_sd = sd(age),
    age_median = median(age),
    age_iqr = IQR(age),
    dependents_mean = mean(dependents),
    dependents_sd = sd(dependents),
    distance_mean = mean(distance_to_provider),
    distance_sd = sd(distance_to_provider),
    distance_median = median(distance_to_provider),
    distance_iqr = IQR(distance_to_provider)
  )
```

### Categorical Variables

```{r}
#| label: categorical-summary

freq_table <- function(data, var) {
  data |>
    count({{ var }}, name = "n") |>
    mutate(pct = 100 * n / sum(n)) |>
    arrange(desc(n))
}

freq_table(analysis_df, self_report_income)
freq_table(analysis_df, edu)
freq_table(analysis_df, race_ethnicity)
freq_table(analysis_df, insurance)
freq_table(analysis_df, job_type)
```

### Clinical Risk Factors

```{r}
#| label: risk-factor-summary

analysis_df |>
  summarise(across(all_of(binary_vars), ~ mean(.))) |>
  pivot_longer(everything(), names_to = "risk_factor", values_to = "prevalence") |>
  mutate(prevalence_pct = 100 * prevalence) |>
  select(risk_factor, prevalence_pct) |>
  arrange(desc(prevalence_pct))
```

### Outcome by Key Groups

```{r}
#| label: outcome-by-groups

analysis_df |>
  group_by(insurance) |>
  summarise(
    n = n(),
    postnatal_care_rate = mean(received_comprehensive_postnatal_care),
    .groups = "drop"
  ) |>
  arrange(desc(postnatal_care_rate))
```

```{r}
#| label: outcome-by-income

analysis_df |>
  group_by(self_report_income) |>
  summarise(
    n = n(),
    postnatal_care_rate = mean(received_comprehensive_postnatal_care),
    .groups = "drop"
  ) |>
  arrange(desc(postnatal_care_rate))
```

## Modeling Strategy

We model `received_comprehensive_postnatal_care` as a binary outcome using logistic regression. This is a good fit because:

1. The target is binary.
2. Coefficients are easy to interpret as odds ratios for policy and clinical audiences.
3. We can include both social and clinical covariates in one model.

To capture potential nonlinearity, age and distance to provider are modeled with natural splines.

## Model Fitting and Performance

```{r}
#| label: train-test-split

set.seed(2026)
idx <- sample.int(nrow(analysis_df), size = floor(0.8 * nrow(analysis_df)))

train_df <- analysis_df[idx, ]
test_df <- analysis_df[-idx, ]

model_formula <- received_comprehensive_postnatal_care ~
  ns(age, df = 4) +
  self_report_income +
  edu +
  race_ethnicity +
  insurance +
  job_type +
  dependents +
  ns(distance_to_provider, df = 4) +
  obesity +
  multiple_gestation +
  diabetes +
  heart_disease +
  placenta_previa +
  hypertension +
  gest_hypertension +
  preeclampsia +
  state

logit_fit <- glm(model_formula, data = train_df, family = binomial())
```

```{r}
#| label: model-performance

test_prob <- predict(logit_fit, newdata = test_df, type = "response")
test_y <- test_df$received_comprehensive_postnatal_care

test_auc <- auc_fast(test_y, test_prob)
threshold <- mean(train_df$received_comprehensive_postnatal_care)
test_pred <- as.integer(test_prob >= threshold)

tibble(
  metric = c("AUC", "Threshold (train prevalence)"),
  value = c(test_auc, threshold)
)
```

```{r}
#| label: model-classification

classification_metrics(test_y, test_pred)
```

## Model Report

```{r}
#| label: model-odds-ratios

coef_tbl <- summary(logit_fit)$coefficients

or_tbl <- tibble(
  term = rownames(coef_tbl),
  estimate = coef_tbl[, "Estimate"],
  std_error = coef_tbl[, "Std. Error"],
  z_value = coef_tbl[, "z value"],
  p_value = coef_tbl[, "Pr(>|z|)"],
  odds_ratio = exp(estimate),
  ci_low = exp(estimate - 1.96 * std_error),
  ci_high = exp(estimate + 1.96 * std_error)
) |>
  arrange(p_value)

or_tbl
```

```{r}
#| label: top-effects

or_tbl |>
  filter(term != "(Intercept)") |>
  mutate(abs_log_or = abs(log(odds_ratio))) |>
  arrange(desc(abs_log_or)) |>
  select(term, odds_ratio, ci_low, ci_high, p_value) |>
  slice_head(n = 15)
```
