---
title: "Tools Comparison (Detailed): Codex CLI + Codex GPT-5.2"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `codex-cli`
- Model: `codex-gpt-5-2`
- Prompt type: `detailed`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## Prompt

```text
I want you to help me analyze this dataset: `data/synthetic/simulated_maternal_health_data.csv`

My goal is to produce a thorough analysis that is statistically sound and lays out all steps of the analysis.

Please do the following:

1. Load the dataset and normalize columns/types.
   - Use clear snake_case names.
   - Ensure categorical vs numeric types are explicit.

2. Summarize the data.
   - Report row count, column count, missingness, and key descriptive summaries.
   - Flag suspicious values or quality concerns.

3. Create key visualizations.
   - Show relationships for:
     - `received_comprehensive_postnatal_care`
     - `distance_to_provider`
     - `insurance`
     - `race_ethnicity`

4. Build a count outcome and model it.
   - Define `comorbidity_count` as the row-wise sum of:
     - `obesity`,
     - `multiple_gestation`,
     - `diabetes`,
     - `heart_disease`,
     - `placenta_previa`,
     - `hypertension`,
     - `gest_hypertension`,
     - `preeclampsia`.
   - Select an appropriate count-model strategy (Poisson vs negative binomial, with justification).
   - Fit the model and report interpretable results.

5. Report results responsibly.
   - Do not overstate causality.
   - Include model assumptions, diagnostics, and limitations.

Constraints:
- Prefer readable, modular, reproducible code that follows R community idioms.
- Report back a summary of what you changed.
```

## Load And Normalize

```{r}
#| label: load-normalize
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(MASS)

raw_df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))

normalize_names <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9]+", "_", x)
  x <- gsub("^_|_$", "", x)
  x
}

df <- raw_df
names(df) <- normalize_names(names(df))

categorical_vars <- c(
  "state",
  "self_report_income",
  "edu",
  "race_ethnicity",
  "insurance",
  "job_type",
  "received_comprehensive_postnatal_care"
)

numeric_vars <- c(
  "id",
  "provider_id",
  "age",
  "dependents",
  "distance_to_provider",
  "obesity",
  "multiple_gestation",
  "diabetes",
  "heart_disease",
  "placenta_previa",
  "hypertension",
  "gest_hypertension",
  "preeclampsia"
)

df <- df %>%
  mutate(across(all_of(categorical_vars), as.factor)) %>%
  mutate(across(all_of(numeric_vars), as.numeric))
```

## Data Summary

```{r}
#| label: data-summary

row_count <- nrow(df)
col_count <- ncol(df)

missing_summary <- df %>%
  summarize(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing")

summary_numeric <- df %>%
  summarize(across(where(is.numeric), list(
    min = ~ min(.x, na.rm = TRUE),
    median = ~ median(.x, na.rm = TRUE),
    mean = ~ mean(.x, na.rm = TRUE),
    max = ~ max(.x, na.rm = TRUE)
  )))

row_count
col_count
missing_summary
summary_numeric
```

```{r}
#| label: suspicious-values

suspicious <- list(
  age_out_of_range = df %>% filter(age < 10 | age > 55) %>% nrow(),
  negative_distance = df %>% filter(distance_to_provider < 0) %>% nrow(),
  insurance_unknown = df %>% filter(is.na(insurance)) %>% nrow()
)

suspicious
```

## Key Visualizations

```{r}
#| label: viz-postnatal-care

ggplot(df, aes(x = received_comprehensive_postnatal_care)) +
  geom_bar(fill = "#3a5a40") +
  labs(
    title = "Comprehensive Postnatal Care",
    x = "Received Comprehensive Postnatal Care",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
#| label: viz-distance

ggplot(df, aes(x = distance_to_provider)) +
  geom_histogram(bins = 30, fill = "#2a9d8f") +
  labs(
    title = "Distance To Provider",
    x = "Distance",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
#| label: viz-insurance

ggplot(df, aes(x = insurance)) +
  geom_bar(fill = "#e76f51") +
  labs(
    title = "Insurance Distribution",
    x = "Insurance",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
#| label: viz-race-ethnicity

ggplot(df, aes(x = race_ethnicity)) +
  geom_bar(fill = "#264653") +
  labs(
    title = "Race/Ethnicity Distribution",
    x = "Race/Ethnicity",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Count Outcome Modeling

```{r}
#| label: comorbidity-count

comorbidity_vars <- c(
  "obesity",
  "multiple_gestation",
  "diabetes",
  "heart_disease",
  "placenta_previa",
  "hypertension",
  "gest_hypertension",
  "preeclampsia"
)

df <- df %>%
  mutate(comorbidity_count = rowSums(across(all_of(comorbidity_vars)), na.rm = TRUE))

count_summary <- df %>%
  summarize(
    mean_count = mean(comorbidity_count),
    var_count = var(comorbidity_count)
  )

count_summary
```

```{r}
#| label: count-model

poisson_fit <- glm(
  comorbidity_count ~ age + dependents + distance_to_provider +
    insurance + race_ethnicity + received_comprehensive_postnatal_care,
  data = df,
  family = poisson()
)

overdispersion_ratio <- sum(residuals(poisson_fit, type = "pearson")^2) /
  poisson_fit$df.residual

use_negbin <- overdispersion_ratio > 1.2

if (use_negbin) {
  count_fit <- MASS::glm.nb(
    comorbidity_count ~ age + dependents + distance_to_provider +
      insurance + race_ethnicity + received_comprehensive_postnatal_care,
    data = df
  )
  model_choice <- "Negative Binomial (overdispersion detected)"
} else {
  count_fit <- poisson_fit
  model_choice <- "Poisson (no strong overdispersion detected)"
}

model_choice
overdispersion_ratio
summary(count_fit)
```

```{r}
#| label: count-model-results

coefs <- coef(count_fit)
conf <- suppressMessages(confint(count_fit))

results <- data.frame(
  term = names(coefs),
  estimate = exp(coefs),
  conf_low = exp(conf[, 1]),
  conf_high = exp(conf[, 2])
)

results
```

## Model Assumptions, Diagnostics, And Limitations

```{r}
#| label: diagnostics

par(mfrow = c(2, 2))
plot(count_fit)
par(mfrow = c(1, 1))
```

Key considerations:

- The outcome is a sum of binary conditions, so a count model is appropriate.
- If overdispersion is present (variance > mean), a negative binomial model is preferred.
- Coefficients are interpreted as multiplicative effects on the expected count.
- Associations do not imply causality; this is observational synthetic data.
- Diagnostics focus on residual patterns and dispersion, but do not establish causal validity.

## Summary Of Changes

- Added data normalization (snake_case names, explicit categorical and numeric types).
- Added missingness and descriptive summaries plus basic data quality flags.
- Built key visualizations for postnatal care, distance to provider, insurance, and race/ethnicity.
- Created `comorbidity_count`, selected Poisson vs negative binomial based on dispersion, and fit the model.
- Added model diagnostics and responsible reporting guidance.
