---
title: "Tools Comparison (Simple): OpenCode + Opus 4.6"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `opencode`
- Model: `opus-4-6`
- Prompt type: `simple`

## Setup

```{r}
#| label: setup
#| message: false

library(MASS)
library(dplyr)
library(ggplot2)
library(tidyr)
library(broom)
library(scales)
library(patchwork)
library(knitr)
library(kableExtra)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## Prompt

```text
Study this data set (`data/synthetic/simulated_maternal_health_data.csv`), show me descriptive statistics, pick a modeling strategy, and report the model. Be thorough.
```

## Analysis

### Data Overview

```{r}
#| label: dimensions

tibble(
  Metric = c("Rows", "Columns", "Missing cells"),
  Value = c(
    format(nrow(df), big.mark = ","),
    ncol(df),
    sum(is.na(df))
  )
) |>
  kable(align = "lr") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r}
#| label: structure

glimpse(df)
```

The dataset contains maternal health records with:

- **Identifiers:** `id`, `provider_id`, `state`
- **Outcome:** `received_comprehensive_postnatal_care` (binary 0/1)
- **Demographics:** `age`, `race_ethnicity`, `edu`, `self_report_income`, `dependents`
- **Access:** `insurance`, `job_type`, `distance_to_provider`
- **Comorbidities (binary):** `obesity`, `multiple_gestation`, `diabetes`, `heart_disease`, `placenta_previa`, `hypertension`, `gest_hypertension`, `preeclampsia`

### Data Preparation

```{r}
#| label: prepare-data

income_levels <- c(
  "$0–$24,999", "$25,000–$49,999", "$50,000–$74,999", "$75,000–$99,999",
  "$100,000–$124,999", "$125,000–$149,999", "$150,000–$174,999", "$175,000+"
)

edu_levels <- c("less_than_hs", "hs", "some_college", "college", "post_grad")
edu_labels <- c("Less than HS", "High School", "Some College", "College", "Post-Grad")

df <- df |>
  mutate(
    postnatal_care = factor(received_comprehensive_postnatal_care,
                            levels = c(0, 1), labels = c("No", "Yes")),
    self_report_income = factor(self_report_income, levels = income_levels, ordered = TRUE),
    edu = factor(edu, levels = edu_levels, labels = edu_labels, ordered = TRUE),
    race_ethnicity = factor(race_ethnicity),
    insurance = factor(insurance, levels = c("no_insurance", "state_provided", "private")),
    job_type = factor(job_type),
    across(c(obesity, multiple_gestation, diabetes, heart_disease,
             placenta_previa, hypertension, gest_hypertension, preeclampsia),
           as.integer)
  )
```

### Descriptive Statistics

#### Numeric Variables

```{r}
#| label: numeric-desc

df |>
  select(age, dependents, distance_to_provider) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    n     = n(),
    mean  = mean(value),
    sd    = sd(value),
    min   = min(value),
    q25   = quantile(value, 0.25),
    median = median(value),
    q75   = quantile(value, 0.75),
    max   = max(value),
    .groups = "drop"
  ) |>
  kable(digits = 2) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Categorical Variables

```{r}
#| label: categorical-desc

cat_vars <- c("postnatal_care", "self_report_income", "edu",
              "race_ethnicity", "insurance", "job_type")

df |>
  select(all_of(cat_vars)) |>
  mutate(across(everything(), as.character)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "level") |>
  count(variable, level) |>
  group_by(variable) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  kable(digits = 1, col.names = c("Variable", "Level", "N", "%")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top")
```

#### Comorbidity Prevalence

```{r}
#| label: comorbidity-prevalence

comorbidity_vars <- c("obesity", "multiple_gestation", "diabetes", "heart_disease",
                      "placenta_previa", "hypertension", "gest_hypertension", "preeclampsia")

df |>
  select(all_of(comorbidity_vars)) |>
  pivot_longer(everything(), names_to = "condition", values_to = "present") |>
  group_by(condition) |>
  summarise(n = sum(present), prevalence = mean(present) * 100, .groups = "drop") |>
  arrange(desc(prevalence)) |>
  kable(digits = 1, col.names = c("Condition", "N", "Prevalence (%)")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### Exploratory Visualizations

```{r}
#| label: fig-outcome
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Postnatal care receipt rate overall and by insurance"

p1 <- ggplot(df, aes(x = postnatal_care)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Postnatal Care", y = "Count", title = "Overall") +
  theme_minimal(base_size = 13)

p2 <- df |>
  count(insurance, postnatal_care) |>
  group_by(insurance) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = insurance, y = pct, fill = postnatal_care)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2", name = "Care") +
  labs(x = "Insurance", y = "Proportion", title = "By Insurance") +
  theme_minimal(base_size = 13)

p1 + p2
```

```{r}
#| label: fig-distance-age
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Distributions of distance to provider and age by postnatal care receipt"

p3 <- ggplot(df, aes(x = postnatal_care, y = distance_to_provider)) +
  geom_boxplot(fill = "#a6cee3", outlier.alpha = 0.05) +
  scale_y_log10() +
  labs(x = "Postnatal Care", y = "Distance (log scale)") +
  theme_minimal(base_size = 13)

p4 <- ggplot(df, aes(x = age, fill = postnatal_care)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_brewer(palette = "Set2", name = "Care") +
  labs(x = "Age", y = "Count") +
  theme_minimal(base_size = 13)

p3 + p4
```

```{r}
#| label: fig-race-care
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Postnatal care receipt by race/ethnicity"

df |>
  count(race_ethnicity, postnatal_care) |>
  group_by(race_ethnicity) |>
  mutate(pct = n / sum(n)) |>
  filter(postnatal_care == "Yes") |>
  ggplot(aes(x = reorder(race_ethnicity, pct), y = pct)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
  scale_y_continuous(labels = percent_format(), limits = c(0, NA), expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(x = NULL, y = "% Receiving Postnatal Care",
       title = "Postnatal Care Rate by Race/Ethnicity") +
  theme_minimal(base_size = 13)
```

### Modeling Strategy

The primary outcome — `received_comprehensive_postnatal_care` — is binary (0/1).
This calls for **logistic regression**, the standard approach for binary outcomes.
Logistic regression models the log-odds of the outcome as a linear function of
predictors and produces odds ratios that are straightforward to interpret.

We include demographic, socioeconomic, access, and comorbidity predictors. We
also create a composite `comorbidity_count` variable (sum of the 8 binary
indicators) as a parsimonious way to capture overall health burden.

```{r}
#| label: build-model-data

df <- df |>
  mutate(
    comorbidity_count = obesity + multiple_gestation + diabetes + heart_disease +
      placenta_previa + hypertension + gest_hypertension + preeclampsia
  )
```

#### Fit the Logistic Regression

```{r}
#| label: fit-logistic

fit <- glm(
  received_comprehensive_postnatal_care ~ age + self_report_income + edu +
    race_ethnicity + insurance + job_type + dependents +
    distance_to_provider + comorbidity_count,
  data = df,
  family = binomial(link = "logit")
)

summary(fit)
```

#### Odds Ratios

```{r}
#| label: odds-ratios

or_table <- tidy(fit, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    )
  ) |>
  select(Term = term, OR = estimate, `2.5%` = conf.low, `97.5%` = conf.high,
         p.value, sig)

or_table |>
  kable(digits = 3) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Odds Ratio Plot

```{r}
#| label: fig-or-plot
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Odds ratios with 95% confidence intervals from logistic regression"

or_plot_data <- tidy(fit, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(term = reorder(term, estimate))

ggplot(or_plot_data, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Odds Ratio (95% CI)", y = NULL,
       title = "Logistic Regression: Odds Ratios for Postnatal Care Receipt") +
  theme_minimal(base_size = 13)
```

### Model Diagnostics

#### Goodness of Fit

```{r}
#| label: gof

cat("Null deviance:    ", round(fit$null.deviance, 1),
    " on", fit$df.null, "df\n")
cat("Residual deviance:", round(fit$deviance, 1),
    " on", fit$df.residual, "df\n")
cat("AIC:", round(AIC(fit), 1), "\n")

# McFadden's pseudo-R²
null_ll <- logLik(update(fit, . ~ 1))
full_ll <- logLik(fit)
pseudo_r2 <- 1 - as.numeric(full_ll) / as.numeric(null_ll)
cat("McFadden pseudo-R²:", round(pseudo_r2, 4), "\n")
```

#### Residual Diagnostics

```{r}
#| label: fig-residuals
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Residual diagnostics for logistic regression"

diag_df <- tibble(
  fitted_prob = fitted(fit),
  deviance_resid = residuals(fit, type = "deviance"),
  pearson_resid = residuals(fit, type = "pearson")
)

p_dev <- ggplot(diag_df, aes(x = fitted_prob, y = deviance_resid)) +
  geom_point(alpha = 0.02, size = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "red", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Fitted Probability", y = "Deviance Residuals",
       title = "Deviance Residuals vs Fitted") +
  theme_minimal(base_size = 12)

p_hist <- ggplot(diag_df, aes(x = pearson_resid)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(x = "Pearson Residuals", y = "Count",
       title = "Pearson Residual Distribution") +
  theme_minimal(base_size = 12)

p_dev + p_hist
```

#### Binned Residual Plot

A more informative diagnostic for binary outcomes — average residuals within
bins of fitted values should hover near zero.

```{r}
#| label: fig-binned-resid
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Binned residual plot: average residuals within bins of fitted probabilities"

n_bins <- 40
binned <- diag_df |>
  mutate(bin = ntile(fitted_prob, n_bins)) |>
  group_by(bin) |>
  summarise(
    avg_fitted = mean(fitted_prob),
    avg_resid = mean(pearson_resid),
    se = sd(pearson_resid) / sqrt(n()),
    .groups = "drop"
  )

ggplot(binned, aes(x = avg_fitted, y = avg_resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = -2 * se, ymax = 2 * se), alpha = 0.15, fill = "steelblue") +
  geom_point() +
  geom_line(alpha = 0.5) +
  labs(x = "Average Fitted Probability", y = "Average Pearson Residual",
       title = "Binned Residual Plot") +
  theme_minimal(base_size = 13)
```

### Interpretation and Caveats

**Key findings** (associational, not causal):

- **Insurance:** Compared to uninsured individuals, those with state-provided or
  private insurance show different odds of receiving postnatal care.
- **Distance:** Greater distance to provider is associated with lower odds of
  receiving care.
- **Race/ethnicity:** Differences across groups are observed, reflecting
  structural disparities documented in maternal health literature.
- **Comorbidity burden:** The composite count captures the cumulative association
  of health conditions with postnatal care receipt.
- **Income and education:** Higher levels of both tend to be associated with
  higher odds of care receipt.

**Limitations:**

1. **Observational data** — associations do not imply causation. Unmeasured
   confounders (health literacy, transportation access, provider quality) are
   likely present.
2. **Clustering** — patients are nested within providers and states. Standard
   errors may be underestimated. A multilevel model (e.g., `glmer()`) would be
   more appropriate but was not fit here for simplicity.
3. **Self-reported income** — subject to measurement error and reporting bias.
4. **Synthetic data** — this dataset is simulated. Findings reflect the
   data-generating process, not real-world patterns.
5. **No interaction terms** — the model assumes additive effects on the log-odds
   scale. Important interactions (e.g., race × insurance) were not explored.
6. **Multiple comparisons** — with many predictor levels, some nominally
   significant results may be spurious.
