---
title: "Tools Comparison (Simple): Codex CLI + Codex GPT-5.2"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `codex-cli`
- Model: `codex-gpt-5-2`
- Prompt type: `simple`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## Prompt

```text
Study this data set (`data/synthetic/simulated_maternal_health_data.csv`), show me descriptive statistics, pick a modeling strategy, and report the model. Be thorough.
```

## Descriptive Statistics

### Data Overview

```{r}
#| label: overview

library(tidyr)
library(knitr)

overview <- tibble::tibble(
  n = nrow(df),
  providers = dplyr::n_distinct(df$provider_id),
  states = dplyr::n_distinct(df$state),
  outcome_rate = mean(df$received_comprehensive_postnatal_care)
)

knitr::kable(overview, digits = 3)
```

### Missingness

```{r}
#| label: missingness

missingness <- tibble::tibble(
  variable = names(df),
  missing_n = sapply(df, function(x) sum(is.na(x)))
)

knitr::kable(missingness)
```

### Numeric Summaries

```{r}
#| label: numeric-summaries

num_vars <- c("age", "dependents", "distance_to_provider")

num_summary <- df |>
  dplyr::select(dplyr::all_of(num_vars)) |>
  dplyr::summarise(
    dplyr::across(
      .cols = dplyr::everything(),
      .fns = list(
        min = min,
        q1 = ~quantile(.x, 0.25),
        median = median,
        mean = mean,
        q3 = ~quantile(.x, 0.75),
        max = max
      )
    )
  ) |>
  tidyr::pivot_longer(cols = dplyr::everything()) |>
  tidyr::separate(name, into = c("variable", "stat"), sep = "_") |>
  tidyr::pivot_wider(names_from = stat, values_from = value)

knitr::kable(num_summary, digits = 3)
```

### Categorical Distributions

```{r}
#| label: categorical-distributions

cat_vars <- c(
  "state",
  "self_report_income",
  "edu",
  "race_ethnicity",
  "insurance",
  "job_type"
)

cat_table <- function(var) {
  df |>
    dplyr::count(.data[[var]]) |>
    dplyr::mutate(pct = n / sum(n)) |>
    dplyr::arrange(dplyr::desc(n)) |>
    dplyr::rename(level = 1) |>
    dplyr::mutate(variable = var) |>
    dplyr::select(variable, level, n, pct)
}

cat_dist <- dplyr::bind_rows(lapply(cat_vars, cat_table))

knitr::kable(cat_dist, digits = 3)
```

### Outcome Rates by Category

```{r}
#| label: outcome-by-category

outcome_by <- function(var) {
  df |>
    dplyr::group_by(.data[[var]]) |>
    dplyr::summarise(
      n = dplyr::n(),
      outcome_rate = mean(received_comprehensive_postnatal_care),
      .groups = "drop"
    ) |>
    dplyr::arrange(dplyr::desc(outcome_rate)) |>
    dplyr::rename(level = 1) |>
    dplyr::mutate(variable = var) |>
    dplyr::select(variable, level, n, outcome_rate)
}

outcome_rates <- dplyr::bind_rows(lapply(cat_vars, outcome_by))

knitr::kable(outcome_rates, digits = 3)
```

### Clinical Prevalence

```{r}
#| label: clinical-prevalence

binary_vars <- c(
  "obesity",
  "multiple_gestation",
  "diabetes",
  "heart_disease",
  "placenta_previa",
  "hypertension",
  "gest_hypertension",
  "preeclampsia"
)

binary_prev <- tibble::tibble(
  variable = binary_vars,
  prevalence = sapply(df[binary_vars], mean)
)

knitr::kable(binary_prev, digits = 3)
```

## Modeling Strategy

The outcome is binary, so I fit a multivariable logistic regression with:
age (and age-squared for nonlinearity), dependents, log-transformed distance to provider,
clinical risk factors, socioeconomic variables, insurance, job type, race/ethnicity, and
state fixed effects. This provides an interpretable baseline model for inference and
adjusted associations.

## Model Results

```{r}
#| label: model

df_model <- df |>
  dplyr::mutate(
    state = factor(state),
    self_report_income = factor(self_report_income),
    edu = factor(edu),
    race_ethnicity = factor(race_ethnicity),
    insurance = factor(insurance),
    job_type = factor(job_type),
    log_distance = log1p(distance_to_provider)
  ) |>
  dplyr::mutate(
    self_report_income = relevel(self_report_income, ref = "$50,000–$74,999"),
    edu = relevel(edu, ref = "hs"),
    race_ethnicity = relevel(race_ethnicity, ref = "white"),
    insurance = relevel(insurance, ref = "no_insurance"),
    job_type = relevel(job_type, ref = "unskilled")
  )

model <- glm(
  received_comprehensive_postnatal_care ~
    age + I(age^2) + dependents + log_distance +
    obesity + multiple_gestation + diabetes + heart_disease +
    placenta_previa + hypertension + gest_hypertension + preeclampsia +
    self_report_income + edu + race_ethnicity + insurance + job_type +
    state,
  data = df_model,
  family = binomial()
)

coefs <- coef(model)
ses <- sqrt(diag(vcov(model)))
or <- exp(coefs)
lcl <- exp(coefs - 1.96 * ses)
ucl <- exp(coefs + 1.96 * ses)

terms_keep <- c(
  "age", "I(age^2)", "dependents", "log_distance",
  "obesity", "multiple_gestation", "diabetes", "heart_disease",
  "placenta_previa", "hypertension", "gest_hypertension", "preeclampsia",
  "self_report_income$25,000–$49,999", "self_report_income$0–$24,999",
  "self_report_income$75,000–$99,999", "self_report_income$100,000–$124,999",
  "self_report_income$125,000–$149,999", "self_report_income$150,000–$174,999",
  "self_report_income$175,000+",
  "educollege", "edupost_grad", "edusome_college",
  "race_ethnicityblack", "race_ethnicityhispanic", "race_ethnicityasian",
  "race_ethnicityaian", "race_ethnicityother", "race_ethnicitynhpi",
  "insuranceprivate", "insurancestate_provided",
  "job_typeprofessional", "job_typetrade", "job_typeunemployed"
)

terms_keep <- terms_keep[terms_keep %in% names(coefs)]

or_table <- tibble::tibble(
  term = terms_keep,
  OR = or[terms_keep],
  LCL = lcl[terms_keep],
  UCL = ucl[terms_keep]
)

knitr::kable(or_table, digits = 3)
```

### Model Fit

```{r}
#| label: model-fit

pred <- predict(model, type = "response")
brier <- mean((pred - df_model$received_comprehensive_postnatal_care)^2)
acc <- mean(ifelse(pred >= 0.5, 1, 0) == df_model$received_comprehensive_postnatal_care)
baseline_acc <- max(
  mean(df_model$received_comprehensive_postnatal_care == 1),
  mean(df_model$received_comprehensive_postnatal_care == 0)
)

fit <- tibble::tibble(
  AIC = AIC(model),
  McFadden_R2 = 1 - (model$deviance / model$null.deviance),
  Brier = brier,
  Accuracy_0_5 = acc,
  Baseline_Accuracy = baseline_acc
)

knitr::kable(fit, digits = 4)
```

## Summary

The model shows modest associations between the outcome and access (distance), socioeconomic
factors (income, insurance, job type), and some clinical conditions, with notable differences
by race/ethnicity and state. Overall fit is limited, suggesting additional variables or
modeling strategies (provider-level effects, interactions, or non-linear models) would be
needed for strong predictive performance.
