---
title: "Tools Comparison (Detailed): OpenCode + Codex GPT-5.3"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false
```

## Run Metadata

- Harness: `opencode`
- Model: `codex-gpt-5-3`
- Prompt type: `detailed`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)

df_raw <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))

normalize_names <- function(x) {
  y <- tolower(x)
  y <- gsub("[^a-z0-9]+", "_", y)
  y <- gsub("_+", "_", y)
  y <- gsub("^_|_$", "", y)
  y
}

df <- df_raw
names(df) <- normalize_names(names(df))

income_levels <- c(
  "<$25,000", "$25,000-$49,999", "$50,000-$74,999",
  "$75,000-$99,999", "$100,000+"
)

education_levels <- c("less_than_hs", "hs", "some_college", "college_plus")

binary_vars <- c(
  "received_comprehensive_postnatal_care",
  "obesity", "multiple_gestation", "diabetes", "heart_disease",
  "placenta_previa", "hypertension", "gest_hypertension", "preeclampsia"
)

df <- df |>
  mutate(
    id = as.integer(id),
    provider_id = as.integer(provider_id),
    state = as.factor(state),
    self_report_income = as.character(self_report_income),
    self_report_income = stringr::str_replace_all(self_report_income, "–", "-"),
    self_report_income = factor(self_report_income, levels = income_levels, ordered = TRUE),
    age = as.integer(age),
    edu = factor(edu, levels = education_levels, ordered = TRUE),
    race_ethnicity = as.factor(race_ethnicity),
    insurance = as.factor(insurance),
    job_type = as.factor(job_type),
    dependents = as.integer(dependents),
    distance_to_provider = as.numeric(distance_to_provider),
    across(all_of(binary_vars), as.integer)
  )
```

## Data Summary and Quality Checks

```{r}
#| label: summary

row_count <- nrow(df)
col_count <- ncol(df)

missingness <- tibble(
  column = names(df),
  missing_n = sapply(df, function(x) sum(is.na(x))),
  missing_pct = round(100 * missing_n / row_count, 2)
) |>
  arrange(desc(missing_n), column)

numeric_summary <- df |>
  select(where(is.numeric)) |>
  summarise(across(everything(), list(
    mean = ~mean(.x, na.rm = TRUE),
    sd = ~sd(.x, na.rm = TRUE),
    min = ~min(.x, na.rm = TRUE),
    p25 = ~quantile(.x, 0.25, na.rm = TRUE),
    median = ~median(.x, na.rm = TRUE),
    p75 = ~quantile(.x, 0.75, na.rm = TRUE),
    max = ~max(.x, na.rm = TRUE)
  ))) |>
  tidyr::pivot_longer(
    cols = everything(),
    names_to = c("variable", "stat"),
    names_sep = "_",
    values_to = "value"
  ) |>
  tidyr::pivot_wider(names_from = stat, values_from = value)

quality_flags <- tibble(
  check = c(
    "duplicate_id_rows",
    "age_outside_12_60",
    "negative_distance",
    "binary_not_0_1"
  ),
  flagged_n = c(
    sum(duplicated(df$id)),
    sum(df$age < 12 | df$age > 60, na.rm = TRUE),
    sum(df$distance_to_provider < 0, na.rm = TRUE),
    sum(!df$received_comprehensive_postnatal_care %in% c(0L, 1L), na.rm = TRUE) +
      sum(!df$obesity %in% c(0L, 1L), na.rm = TRUE) +
      sum(!df$multiple_gestation %in% c(0L, 1L), na.rm = TRUE) +
      sum(!df$diabetes %in% c(0L, 1L), na.rm = TRUE) +
      sum(!df$heart_disease %in% c(0L, 1L), na.rm = TRUE) +
      sum(!df$placenta_previa %in% c(0L, 1L), na.rm = TRUE) +
      sum(!df$hypertension %in% c(0L, 1L), na.rm = TRUE) +
      sum(!df$gest_hypertension %in% c(0L, 1L), na.rm = TRUE) +
      sum(!df$preeclampsia %in% c(0L, 1L), na.rm = TRUE)
  )
)

tibble(metric = c("Rows", "Columns"), value = c(row_count, col_count))
```

```{r}
#| label: missingness
#| tbl-cap: "Missingness by column"

knitr::kable(missingness)
```

```{r}
#| label: numeric-summary
#| tbl-cap: "Numeric descriptive summary"

knitr::kable(numeric_summary, digits = 2)
```

```{r}
#| label: quality-flags
#| tbl-cap: "Basic data quality flags"

knitr::kable(quality_flags)
```

## Key Visualizations

```{r}
#| label: viz-distance-by-care
#| fig-cap: "Distance to provider by postnatal care status"

df |>
  mutate(
    care = factor(
      received_comprehensive_postnatal_care,
      levels = c(0, 1),
      labels = c("No comprehensive care", "Received comprehensive care")
    )
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = care, y = distance_to_provider, fill = care)) +
  ggplot2::geom_boxplot(alpha = 0.75, outlier.alpha = 0.1) +
  ggplot2::coord_cartesian(ylim = quantile(df$distance_to_provider, c(0.01, 0.99), na.rm = TRUE)) +
  ggplot2::guides(fill = "none") +
  ggplot2::labs(
    x = NULL,
    y = "Distance to provider",
    title = "People with and without comprehensive care have different travel-distance distributions"
  )
```

```{r}
#| label: viz-care-by-insurance
#| fig-cap: "Comprehensive postnatal care rate by insurance type"

df |>
  group_by(insurance) |>
  summarise(care_rate = mean(received_comprehensive_postnatal_care, na.rm = TRUE), .groups = "drop") |>
  ggplot2::ggplot(ggplot2::aes(x = reorder(insurance, care_rate), y = care_rate, fill = insurance)) +
  ggplot2::geom_col(alpha = 0.85) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::guides(fill = "none") +
  ggplot2::labs(
    x = "Insurance",
    y = "Comprehensive care rate",
    title = "Care receipt varies meaningfully by insurance"
  ) +
  ggplot2::coord_flip()
```

```{r}
#| label: viz-care-by-race
#| fig-cap: "Comprehensive postnatal care rate by race/ethnicity"

df |>
  group_by(race_ethnicity) |>
  summarise(care_rate = mean(received_comprehensive_postnatal_care, na.rm = TRUE), .groups = "drop") |>
  ggplot2::ggplot(ggplot2::aes(x = reorder(race_ethnicity, care_rate), y = care_rate, fill = race_ethnicity)) +
  ggplot2::geom_col(alpha = 0.85) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::guides(fill = "none") +
  ggplot2::labs(
    x = "Race/ethnicity",
    y = "Comprehensive care rate",
    title = "Observed care rates differ across race/ethnicity groups"
  ) +
  ggplot2::coord_flip()
```

## Count Outcome and Model

```{r}
#| label: model-setup

df_model <- df |>
  mutate(
    comorbidity_count =
      obesity +
      multiple_gestation +
      diabetes +
      heart_disease +
      placenta_previa +
      hypertension +
      gest_hypertension +
      preeclampsia,
    distance_to_provider_10 = distance_to_provider / 10
  )

poisson_fit <- glm(
  comorbidity_count ~
    received_comprehensive_postnatal_care +
    distance_to_provider_10 +
    insurance +
    race_ethnicity +
    age +
    dependents +
    edu +
    self_report_income,
  family = poisson(link = "log"),
  data = df_model
)

poisson_dispersion <- sum(residuals(poisson_fit, type = "pearson")^2) / poisson_fit$df.residual
use_nb <- is.finite(poisson_dispersion) && poisson_dispersion > 1.5

nb_fit <- if (use_nb) MASS::glm.nb(
  comorbidity_count ~
    received_comprehensive_postnatal_care +
    distance_to_provider_10 +
    insurance +
    race_ethnicity +
    age +
    dependents +
    edu +
    self_report_income,
  data = df_model
) else NULL

final_model <- if (use_nb) nb_fit else poisson_fit
model_name <- if (use_nb) "Negative binomial" else "Poisson"

distribution_check <- tibble(
  metric = c(
    "Mean(comorbidity_count)",
    "Var(comorbidity_count)",
    "Poisson overdispersion ratio",
    "Selected model"
  ),
  value = c(
    round(mean(df_model$comorbidity_count, na.rm = TRUE), 3),
    round(var(df_model$comorbidity_count, na.rm = TRUE), 3),
    round(poisson_dispersion, 3),
    model_name
  )
)

coefs <- as.data.frame(summary(final_model)$coefficients)
coefs$term <- rownames(coefs)

stat_col <- if ("z value" %in% names(coefs)) "z value" else "t value"
p_col <- if ("Pr(>|z|)" %in% names(coefs)) "Pr(>|z|)" else "Pr(>|t|)"

results <- coefs |>
  transmute(
    term = term,
    estimate = Estimate,
    std_error = `Std. Error`,
    statistic = .data[[stat_col]],
    p_value = .data[[p_col]],
    irr = exp(estimate),
    irr_low_95 = exp(estimate - 1.96 * std_error),
    irr_high_95 = exp(estimate + 1.96 * std_error)
  )
```

```{r}
#| label: model-selection
#| tbl-cap: "Count-model selection checks"

knitr::kable(distribution_check)
```

```{r}
#| label: model-results
#| tbl-cap: "Exponentiated coefficients (incidence rate ratios)"

knitr::kable(results, digits = 3)
```

## Diagnostics and Limitations

```{r}
#| label: diagnostics
#| fig-cap: "Residual diagnostics for the selected count model"

par(mfrow = c(1, 2))
plot(
  fitted(final_model),
  residuals(final_model, type = "pearson"),
  pch = 16,
  cex = 0.3,
  xlab = "Fitted values",
  ylab = "Pearson residuals",
  main = "Residuals vs fitted"
)
abline(h = 0, lty = 2, col = "red")

qqnorm(residuals(final_model, type = "pearson"), pch = 16, cex = 0.3)
qqline(residuals(final_model, type = "pearson"), col = "red", lty = 2)
par(mfrow = c(1, 1))
```

The model is associational, not causal. Coefficients describe adjusted relationships
between predictors and expected comorbidity counts under the specified model.

Key limitations:

- Potential unmeasured confounding remains (for example, severity and access factors not in the data).
- The count outcome is a constructed sum and may combine clinically heterogeneous conditions.
- Residual diagnostics can detect obvious misspecification, but they cannot guarantee perfect fit.
- Results are specific to this dataset's generating process and may not transport to other populations.

## What Was Changed

- Loaded data with `data_read()` and started the notebook with `scaffold()`.
- Normalized column names and explicitly set categorical/numeric/binary data types.
- Added row/column counts, missingness table, descriptive statistics, and quality flags.
- Added visualizations for comprehensive care vs distance, insurance, and race/ethnicity.
- Created `comorbidity_count`, compared Poisson vs negative binomial using overdispersion, and reported IRRs with confidence intervals.
- Added diagnostics plus responsible interpretation and limitations.
```

## Prompt

```text
I want you to help me analyze this dataset: `data/synthetic/simulated_maternal_health_data.csv`

My goal is to produce a thorough analysis that is statistically sound and lays out all steps of the analysis.

Please do the following:

1. Load the dataset and normalize columns/types.
   - Use clear snake_case names.
   - Ensure categorical vs numeric types are explicit.

2. Summarize the data.
   - Report row count, column count, missingness, and key descriptive summaries.
   - Flag suspicious values or quality concerns.

3. Create key visualizations.
   - Show relationships for:
     - `received_comprehensive_postnatal_care`
     - `distance_to_provider`
     - `insurance`
     - `race_ethnicity`

4. Build a count outcome and model it.
   - Define `comorbidity_count` as the row-wise sum of:
     - `obesity`,
     - `multiple_gestation`,
     - `diabetes`,
     - `heart_disease`,
     - `placenta_previa`,
     - `hypertension`,
     - `gest_hypertension`,
     - `preeclampsia`.
   - Select an appropriate count-model strategy (Poisson vs negative binomial, with justification).
   - Fit the model and report interpretable results.

5. Report results responsibly.
   - Do not overstate causality.
   - Include model assumptions, diagnostics, and limitations.

Constraints:
- Prefer readable, modular, reproducible code that follows R community idioms.
- Report back a summary of what you changed.
```
