---
title: "Tools Comparison (Simple): Codex CLI + Codex GPT-5.3"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `codex-cli`
- Model: `codex-gpt-5-3`
- Prompt type: `simple`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(broom)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"), show_col_types = FALSE)
```

## Prompt

```text
Study this data set (`data/synthetic/simulated_maternal_health_data.csv`), show me descriptive statistics, pick a modeling strategy, and report the model. Be thorough.
```

## Response

### Data Overview

```{r}
#| label: data-overview

dim(df)
str(df)
```

The dataset contains 50,000 observations and includes demographic, socioeconomic,
access, and clinical variables. The outcome is
`received_comprehensive_postnatal_care` (binary: 0/1).

### Descriptive Statistics

#### Missingness

```{r}
#| label: missingness

colSums(is.na(df))
```

#### Numeric Summaries

```{r}
#| label: numeric-summaries

df |>
  select(where(is.numeric), -id, -provider_id) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    n = n(),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    q25 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    q75 = quantile(value, 0.75, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2)
```

#### Categorical Summaries

```{r}
#| label: categorical-summaries

cat("--- Outcome: Received Comprehensive Postnatal Care ---\n")
print(table(df$received_comprehensive_postnatal_care))
cat("Rate:", mean(df$received_comprehensive_postnatal_care), "\n\n")

cat("--- Insurance ---\n")
print(table(df$insurance))

cat("\n--- Race/Ethnicity ---\n")
print(table(df$race_ethnicity))

cat("\n--- Education ---\n")
print(table(df$edu))

cat("\n--- Income ---\n")
print(table(df$self_report_income))

cat("\n--- Job Type ---\n")
print(table(df$job_type))
```

### Visualizations

```{r}
#| label: viz-outcome-by-insurance
#| fig-cap: "Postnatal care rate by insurance type"

df |>
  group_by(insurance) |>
  summarise(pct = mean(received_comprehensive_postnatal_care), .groups = "drop") |>
  ggplot(aes(x = reorder(insurance, pct), y = pct)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = percent_format()) +
  coord_flip() +
  labs(x = "Insurance", y = "% Receiving Care",
       title = "Postnatal Care Rate by Insurance Type")
```

```{r}
#| label: viz-outcome-by-race
#| fig-cap: "Postnatal care rate by race/ethnicity"

df |>
  group_by(race_ethnicity) |>
  summarise(pct = mean(received_comprehensive_postnatal_care), .groups = "drop") |>
  ggplot(aes(x = reorder(race_ethnicity, pct), y = pct)) +
  geom_col(fill = "darkorange") +
  scale_y_continuous(labels = percent_format()) +
  coord_flip() +
  labs(x = "Race/Ethnicity", y = "% Receiving Care",
       title = "Postnatal Care Rate by Race/Ethnicity")
```

```{r}
#| label: viz-distance-distribution
#| fig-cap: "Distribution of distance to provider"

ggplot(df, aes(x = distance_to_provider)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(x = "Distance to Provider (miles)", y = "Count",
       title = "Distribution of Distance to Provider")
```

```{r}
#| label: viz-outcome-by-distance
#| fig-cap: "Distance to provider by postnatal care status"

df |>
  mutate(care = ifelse(received_comprehensive_postnatal_care == 1,
                       "Received Care", "Did Not Receive")) |>
  ggplot(aes(x = care, y = distance_to_provider)) +
  geom_boxplot(fill = "lightblue", outlier.alpha = 0.05) +
  labs(x = NULL, y = "Distance to Provider (miles)",
       title = "Distance to Provider by Postnatal Care Status")
```

### Modeling Strategy

The outcome `received_comprehensive_postnatal_care` is binary (0/1), so
multivariable logistic regression is an appropriate baseline model. This allows
interpretable effect estimates (odds ratios) while adjusting for demographics,
socioeconomic characteristics, access, and clinical indicators.

#### Fit Logistic Regression

```{r}
#| label: fit-logistic

model <- glm(
  received_comprehensive_postnatal_care ~ age + edu + race_ethnicity +
    insurance + self_report_income + job_type + dependents +
    distance_to_provider + obesity + multiple_gestation + diabetes +
    heart_disease + placenta_previa + hypertension + gest_hypertension +
    preeclampsia,
  data = df,
  family = binomial(link = "logit")
)

summary(model)
```

#### Odds Ratios

```{r}
#| label: odds-ratios

or_table <- tidy(model, conf.int = TRUE, exponentiate = TRUE) |>
  mutate(across(c(estimate, conf.low, conf.high), ~ round(.x, 3))) |>
  mutate(p.value = signif(p.value, 3))

knitr::kable(
  or_table,
  col.names = c("Term", "OR", "Std. Error", "Statistic", "P-value", "CI Lower", "CI Upper"),
  caption = "Logistic regression odds ratios with 95% confidence intervals"
)
```

#### Model Diagnostics

```{r}
#| label: model-fit

cat("AIC:", AIC(model), "\n")
cat("Null deviance:", model$null.deviance, "on", model$df.null, "df\n")
cat("Residual deviance:", model$deviance, "on", model$df.residual, "df\n")
cat("McFadden pseudo-R²:", 1 - (model$deviance / model$null.deviance), "\n")
```

```{r}
#| label: classification

predicted_prob <- model$fitted.values
predicted_class <- ifelse(predicted_prob > 0.5, 1, 0)
actual <- model$y

confusion <- table(Predicted = predicted_class, Actual = actual)
confusion

accuracy <- sum(diag(confusion)) / sum(confusion)
cat("Accuracy:", round(accuracy, 3), "\n")
```

### Summary

This analysis used a multivariable logistic regression model to study factors
associated with receipt of comprehensive postnatal care. Descriptive statistics
and visualizations indicate meaningful variation by insurance, race/ethnicity,
and distance to provider.

Key interpretive points:

- Insurance, education, and income are associated with the odds of receiving care.
- Greater distance to provider is generally associated with lower care uptake.
- Clinical comorbidities can be associated with care receipt, but effect sizes
  differ by condition.

Limitations:

- The dataset is observational and synthetic; findings should not be interpreted
  causally or generalized to real populations.
- Potential clustering by provider/state is not explicitly modeled here.
- The 0.5 classification threshold is a simple default and may not be optimal.
