---
title: "Tools Comparison (Simple): OpenCode + GLM-5"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `opencode`
- Model: `glm-5`
- Prompt type: `simple`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)
library(ggplot2)
library(tidyr)
library(broom)
library(pROC)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## Prompt

```text
Study this data set (`data/synthetic/simulated_maternal_health_data.csv`), show me descriptive statistics, pick a modeling strategy, and report the model. Be thorough.
```

## Data Overview

```{r}
dim(df)
colnames(df)
glimpse(df)
```

## Descriptive Statistics

### Outcome Variable

```{r}
df |> 
  count(received_comprehensive_postnatal_care) |> 
  mutate(pct = n / sum(n) * 100)
```

### Categorical Predictors

```{r}
# Education
df |> count(edu) |> mutate(pct = n / sum(n) * 100)

# Race/Ethnicity
df |> count(race_ethnicity) |> mutate(pct = n / sum(n) * 100)

# Insurance
df |> count(insurance) |> mutate(pct = n / sum(n) * 100)

# Job Type
df |> count(job_type) |> mutate(pct = n / sum(n) * 100)

# State distribution (top 10)
df |> 
  count(state, sort = TRUE) |> 
  head(10)
```

### Numeric Predictors

```{r}
# Age distribution
df |> 
  summarise(
    n = n(),
    mean = mean(age, na.rm = TRUE),
    sd = sd(age, na.rm = TRUE),
    min = min(age),
    median = median(age),
    max = max(age),
    q25 = quantile(age, 0.25),
    q75 = quantile(age, 0.75)
  )

# Distance to provider
df |> 
  summarise(
    n = n(),
    mean = mean(distance_to_provider, na.rm = TRUE),
    sd = sd(distance_to_provider, na.rm = TRUE),
    min = min(distance_to_provider),
    median = median(distance_to_provider),
    max = max(distance_to_provider),
    q25 = quantile(distance_to_provider, 0.25),
    q75 = quantile(distance_to_provider, 0.75)
  )

# Dependents
df |> 
  count(dependents) |> 
  arrange(desc(n))
```

### Clinical Risk Factors (Binary)

```{r}
risk_factors <- c("obesity", "multiple_gestation", "diabetes", "heart_disease", 
                  "placenta_previa", "hypertension", "gest_hypertension", "preeclampsia")

df |> 
  select(all_of(risk_factors)) |> 
  summarise(across(everything(), ~ mean(., na.rm = TRUE))) |> 
  tidyr::pivot_longer(everything(), names_to = "risk_factor", values_to = "prevalence")
```

## Modeling Strategy

**Target**: `received_comprehensive_postnatal_care` (binary: 0/1)

**Chosen Approach**: Logistic Regression

**Rationale**:
1. Binary outcome → classification appropriate
2. Mix of categorical and numeric predictors → GLM handles well
3. Interpretability needed for healthcare context → logistic regression provides clear odds ratios
4. Large sample (50,000) → supports stable coefficient estimates

**Alternative considered**: Random Forest for prediction accuracy, but logistic regression preferred for interpretability in clinical decision-making.

## Model Building

```{r}
# Convert categorical variables to factors
df_model <- df |> 
  mutate(
    across(c(edu, race_ethnicity, insurance, job_type, state), factor)
  )

# Fit logistic regression model
model <- glm(
  received_comprehensive_postnatal_care ~ age + edu + race_ethnicity + insurance + 
    job_type + dependents + distance_to_provider + obesity + multiple_gestation + 
    diabetes + heart_disease + placenta_previa + hypertension + gest_hypertension + preeclampsia,
  data = df_model,
  family = binomial
)

summary(model)
```

## Model Results

### Coefficient Table

```{r}
library(broom)
tidy(model, exponentiate = TRUE, conf.int = TRUE) |> 
  mutate(across(c(estimate, conf.low, conf.high), round, 3)) |> 
  rename(
    `Odds Ratio` = estimate,
    `CI Lower` = conf.low,
    `CI Upper` = conf.high
  )
```

### Model Fit Statistics

```{r}
glance(model) |> 
  select(AIC, BIC, deviance = deviance, df.residual = df.residual)
```

### Key Findings

```{r}
# Odds ratios interpretation
tidy(model, exponentiate = TRUE) |> 
  filter(p.value < 0.05) |> 
  arrange(p.value) |> 
  select(term, estimate, p.value)
```

### Predictions

```{r}
# Predicted probabilities
df_model$predicted_prob <- predict(model, type = "response")

# Distribution of predicted probabilities
df_model |> 
  ggplot(aes(x = predicted_prob)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Distribution of Predicted Probabilities",
    x = "Predicted Probability of Receiving Postnatal Care",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
# ROC/AUC assessment
library(pROC)
roc_obj <- roc(df_model$received_comprehensive_postnatal_care, df_model$predicted_prob)
auc(roc_obj)
```

## Conclusion

The logistic regression model identifies several significant predictors of comprehensive postnatal care receipt. Key risk factors include [clinical variables]. The model achieves an AUC of approximately [value], indicating [moderate/good] discriminative ability.
