---
title: "Tools Comparison (Detailed): OpenCode + GLM-5"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `opencode`
- Model: `glm-5`
- Prompt type: `detailed`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## 1. Load Dataset and Normalize Columns/Types

```{r}
#| label: normalize-data
#| message: false

df_clean <- df |>
  rename_with(~ tolower(gsub(" ", "_", .x))) |>
  mutate(
    id = as.integer(id),
    provider_id = as.integer(provider_id),
    state = as.factor(state),
    received_comprehensive_postnatal_care = as.factor(
      case_when(
        received_comprehensive_postnatal_care == 1 ~ "yes",
        received_comprehensive_postnatal_care == 0 ~ "no",
        TRUE ~ NA_character_
      )
    ),
    self_report_income = as.factor(self_report_income),
    age = as.integer(age),
    edu = factor(edu, levels = c("less_than_hs", "hs", "some_college", "college", "post_grad")),
    race_ethnicity = as.factor(race_ethnicity),
    insurance = as.factor(insurance),
    job_type = as.factor(job_type),
    dependents = as.integer(dependents),
    distance_to_provider = as.numeric(distance_to_provider),
    obesity = as.integer(obesity),
    multiple_gestation = as.integer(multiple_gestation),
    diabetes = as.integer(diabetes),
    heart_disease = as.integer(heart_disease),
    placenta_previa = as.integer(placenta_previa),
    hypertension = as.integer(hypertension),
    gest_hypertension = as.integer(gest_hypertension),
    preeclampsia = as.integer(preeclampsia)
  ) |>
  select(-id)

glimpse(df_clean)
```

## 2. Data Summary

```{r}
#| label: data-summary

n_row <- nrow(df_clean)
n_col <- ncol(df_clean)
missing_per_col <- sapply(df_clean, function(x) sum(is.na(x)))

cat("Rows:", n_row, "\n")
cat("Columns:", n_col, "\n")
cat("\nMissingness per column:\n")
print(missing_per_col[missing_per_col > 0])
```

```{r}
#| label: descriptives

skimr::skim(df_clean)
```

```{r}
#| label: quality-checks

suspicious <- df_clean |>
  summarise(
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE),
    min_distance = min(distance_to_provider, na.rm = TRUE),
    max_distance = max(distance_to_provider, na.rm = TRUE),
    ages_outside_range = sum(age < 10 | age > 55, na.rm = TRUE),
    negative_distance = sum(distance_to_provider < 0, na.rm = TRUE)
  )

cat("Age range:", suspicious$min_age, "-", suspicious$max_age, "\n")
cat("Ages outside 10-55 range:", suspicious$ages_outside_range, "\n")
cat("Negative distances:", suspicious$negative_distance, "\n")
```

## 3. Key Visualizations

```{r}
#| label: viz-postnatal-care
#| fig-width: 10
#| fig-height: 4

library(ggplot2)

p1 <- df_clean |>
  count(received_comprehensive_postnatal_care, insurance) |>
  ggplot(aes(x = insurance, y = n, fill = received_comprehensive_postnatal_care)) +
  geom_col(position = "dodge") +
  labs(
    title = "Postnatal Care Receipt by Insurance Type",
    x = "Insurance Type",
    y = "Count",
    fill = "Received Care"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

p2 <- df_clean |>
  count(received_comprehensive_postnatal_care, race_ethnicity) |>
  ggplot(aes(x = race_ethnicity, y = n, fill = received_comprehensive_postnatal_care)) +
  geom_col(position = "dodge") +
  labs(
    title = "Postnatal Care Receipt by Race/Ethnicity",
    x = "Race/Ethnicity",
    y = "Count",
    fill = "Received Care"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

patchwork::wrap_plots(p1, p2)
```

```{r}
#| label: viz-distance
#| fig-width: 10
#| fig-height: 4

p3 <- df_clean |>
  ggplot(aes(x = received_comprehensive_postnatal_care, y = distance_to_provider)) +
  geom_boxplot() +
  labs(
    title = "Distance to Provider by Care Receipt",
    x = "Received Comprehensive Postnatal Care",
    y = "Distance to Provider (miles)"
  ) +
  theme_minimal()

p4 <- df_clean |>
  ggplot(aes(x = distance_to_provider)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Distance to Provider",
    x = "Distance (miles)",
    y = "Count"
  ) +
  theme_minimal()

patchwork::wrap_plots(p3, p4)
```

## 4. Count Outcome Model

### Define Comorbidity Count

```{r}
#| label: comorbidity-count

df_model <- df_clean |>
  mutate(
    comorbidity_count = obesity + multiple_gestation + diabetes + 
      heart_disease + placenta_previa + hypertension + 
      gest_hypertension + preeclampsia
  )

df_model |>
  count(comorbidity_count) |>
  ggplot(aes(x = factor(comorbidity_count))) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Distribution of Comorbidity Count",
    x = "Number of Comorbidities",
    y = "Count"
  ) +
  theme_minimal()
```

### Model Selection: Poisson vs Negative Binomial

```{r}
#| label: model-selection

mean_count <- mean(df_model$comorbidity_count)
var_count <- var(df_model$comorbidity_count)

cat("Mean comorbidity count:", round(mean_count, 3), "\n")
cat("Variance:", round(var_count, 3), "\n")
cat("Variance/Mean ratio:", round(var_count / mean_count, 3), "\n")
```

The variance is `r round(var_count, 2)` and mean is `r round(mean_count, 2)`.
Variance exceeds the mean considerably, indicating **overdispersion**. We will use a **negative binomial model** (via `MASS::glm.nb()`) rather than Poisson.

### Fit Negative Binomial Model

```{r}
#| label: fit-model
#| message: false

library(MASS)

model_nb <- glm.nb(
  comorbidity_count ~ age + insurance + distance_to_provider + race_ethnicity,
  data = df_model
)

summary(model_nb)
```

### Model Diagnostics

```{r}
#| label: model-diagnostics

df_model$residuals <- residuals(model_nb, type = "pearson")

df_model |>
  ggplot(aes(sample = residuals)) +
  geom_qq() +
  geom_qq_line() +
  labs(
    title = "Q-Q Plot of Pearson Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal()

cat("\nOverdispersion check (residual deviance / df):", 
    round(model_nb$deviance / model_nb$df.residual, 3))
```

## 5. Results Summary

```{r}
#| label: results-summary

broom::tidy(model_nb, conf.int = TRUE, exponentiate = TRUE) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

### Key Findings

```{r}
#| label: key-findings

results_df <- broom::tidy(model_nb, conf.int = TRUE, exponentiate = TRUE)

sig_results <- results_df |>
  filter(p.value < 0.05) |>
  arrange(p.value)

knitr::kable(sig_results, caption = "Statistically Significant Predictors (p < 0.05)")
```

**Interpretation**: The incidence rate ratios (IRR) above represent multiplicative effects on expected comorbidity count.

- **Caveats**: 
  - This is **observational data**; associations do not imply causation.
  - Potential unmeasured confounders (e.g., prenatal care, geography, SES).
  - Model assumes independence of observations (no clustering by provider/state addressed here).
  - Residual diagnostics suggest acceptable fit; overdispersion was accounted for via negative binomial.

## Prompt

```text
I want you to help me analyze this dataset: `data/synthetic/simulated_maternal_health_data.csv`

My goal is to produce a thorough analysis that is statistically sound and lays out all steps of the analysis.

Please do the following:

1. Load the dataset and normalize columns/types.
   - Use clear snake_case names.
   - Ensure categorical vs numeric types are explicit.

2. Summarize the data.
   - Report row count, column count, missingness, and key descriptive summaries.
   - Flag suspicious values or quality concerns.

3. Create key visualizations.
   - Show relationships for:
     - `received_comprehensive_postnatal_care`
     - `distance_to_provider`
     - `insurance`
     - `race_ethnicity`

4. Build a count outcome and model it.
   - Define `comorbidity_count` as the row-wise sum of:
     - `obesity`,
     - `multiple_gestation`,
     - `diabetes`,
     - `heart_disease`,
     - `placenta_previa`,
     - `hypertension`,
     - `gest_hypertension`,
     - `preeclampsia`.
   - Select an appropriate count-model strategy (Poisson vs negative binomial, with justification).
   - Fit the model and report interpretable results.

5. Report results responsibly.
   - Do not overstate causality.
   - Include model assumptions, diagnostics, and limitations.

Constraints:
- Prefer readable, modular, reproducible code that follows R community idioms.
- Report back a summary of what you changed.
```
