---
title: "Tools Comparison (Detailed): Codex CLI + Codex GPT-5.3"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `codex-cli`
- Model: `codex-gpt-5-3`
- Prompt type: `detailed`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)
library(ggplot2)
library(readr)
library(stringr)
library(forcats)
library(tidyr)
library(purrr)
library(broom)
library(MASS)

df_raw <- read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"), show_col_types = FALSE)
```

## Prompt

```text
I want you to help me analyze this dataset: `data/synthetic/simulated_maternal_health_data.csv`

My goal is to produce a thorough analysis that is statistically sound and lays out all steps of the analysis.

Please do the following:

1. Load the dataset and normalize columns/types.
   - Use clear snake_case names.
   - Ensure categorical vs numeric types are explicit.

2. Summarize the data.
   - Report row count, column count, missingness, and key descriptive summaries.
   - Flag suspicious values or quality concerns.

3. Create key visualizations.
   - Show relationships for:
     - `received_comprehensive_postnatal_care`
     - `distance_to_provider`
     - `insurance`
     - `race_ethnicity`

4. Build a count outcome and model it.
   - Define `comorbidity_count` as the row-wise sum of:
     - `obesity`,
     - `multiple_gestation`,
     - `diabetes`,
     - `heart_disease`,
     - `placenta_previa`,
     - `hypertension`,
     - `gest_hypertension`,
     - `preeclampsia`.
   - Select an appropriate count-model strategy (Poisson vs negative binomial, with justification).
   - Fit the model and report interpretable results.

5. Report results responsibly.
   - Do not overstate causality.
   - Include model assumptions, diagnostics, and limitations.

Constraints:
- Prefer readable, modular, reproducible code that follows R community idioms.
- Report back a summary of what you changed.
```

## 1) Normalize Columns and Types

```{r}
#| label: normalize

binary_cols <- c(
  "received_comprehensive_postnatal_care",
  "obesity",
  "multiple_gestation",
  "diabetes",
  "heart_disease",
  "placenta_previa",
  "hypertension",
  "gest_hypertension",
  "preeclampsia"
)

factor_cols <- c("state", "edu", "race_ethnicity", "insurance", "job_type", "self_report_income")

df <- df_raw %>%
  rename_with(~ str_replace_all(.x, "[^A-Za-z0-9]+", "_")) %>%
  rename_with(~ str_to_lower(.x)) %>%
  mutate(
    across(all_of(factor_cols), as.factor),
    self_report_income = fct_relevel(
      self_report_income,
      "$0–$24,999",
      "$25,000–$49,999",
      "$50,000–$74,999",
      "$75,000–$99,999",
      "$100,000+"
    ),
    across(all_of(binary_cols), ~ as.integer(.x))
  )

glimpse(df)
```

## 2) Data Summary and Quality Checks

```{r}
#| label: dimensions-missingness

n_rows <- nrow(df)
n_cols <- ncol(df)

missing_summary <- df %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_n") %>%
  mutate(missing_pct = missing_n / n_rows)

list(
  row_count = n_rows,
  column_count = n_cols,
  total_missing = sum(missing_summary$missing_n)
)

missing_summary %>% arrange(desc(missing_n)) %>% print(n = n_cols)
```

```{r}
#| label: descriptive-summary

numeric_summary <- df %>%
  dplyr::select(where(is.numeric)) %>%
  summarise(across(
    everything(),
    list(
      mean = ~ mean(.x, na.rm = TRUE),
      sd = ~ sd(.x, na.rm = TRUE),
      p25 = ~ quantile(.x, 0.25, na.rm = TRUE),
      median = ~ median(.x, na.rm = TRUE),
      p75 = ~ quantile(.x, 0.75, na.rm = TRUE)
    )
  )) %>%
  pivot_longer(everything(), names_to = c("variable", ".value"), names_sep = "_")

categorical_summary <- df %>%
  dplyr::select(where(is.factor)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "level") %>%
  count(variable, level, sort = TRUE) %>%
  group_by(variable) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

numeric_summary
categorical_summary %>% print(n = 60)
```

```{r}
#| label: quality-checks

binary_violations <- map_dfr(binary_cols, function(col_nm) {
  bad_n <- df %>% filter(!(.data[[col_nm]] %in% c(0L, 1L) | is.na(.data[[col_nm]]))) %>% nrow()
  tibble(variable = col_nm, bad_values = bad_n)
})

quality_flags <- tibble(
  check = c(
    "duplicate id rows",
    "age outside [12, 55]",
    "negative distance_to_provider",
    "distance_to_provider > 200"
  ),
  flagged_n = c(
    sum(duplicated(df$id)),
    sum(df$age < 12 | df$age > 55, na.rm = TRUE),
    sum(df$distance_to_provider < 0, na.rm = TRUE),
    sum(df$distance_to_provider > 200, na.rm = TRUE)
  )
)

binary_violations
quality_flags
```

## 3) Key Visualizations

```{r}
#| label: viz-care-by-insurance
#| fig-cap: "Comprehensive postnatal care rates by insurance type"
#| fig-width: 8
#| fig-height: 4.5

df %>%
  group_by(insurance) %>%
  summarise(care_rate = mean(received_comprehensive_postnatal_care, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = fct_reorder(insurance, care_rate), y = care_rate)) +
  geom_col(fill = "#2C7FB8") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(x = "Insurance", y = "Rate receiving comprehensive postnatal care") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: viz-care-by-race
#| fig-cap: "Comprehensive postnatal care rates by race/ethnicity"
#| fig-width: 8
#| fig-height: 4.5

df %>%
  group_by(race_ethnicity) %>%
  summarise(care_rate = mean(received_comprehensive_postnatal_care, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = fct_reorder(race_ethnicity, care_rate), y = care_rate)) +
  geom_col(fill = "#238B45") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(x = "Race/Ethnicity", y = "Rate receiving comprehensive postnatal care") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: viz-distance-distribution
#| fig-cap: "Distance to provider by postnatal care status"
#| fig-width: 8
#| fig-height: 4.5

df %>%
  mutate(care_status = factor(received_comprehensive_postnatal_care, levels = c(0, 1), labels = c("No", "Yes"))) %>%
  ggplot(aes(x = care_status, y = distance_to_provider, fill = care_status)) +
  geom_boxplot(outlier.alpha = 0.15, width = 0.6) +
  scale_fill_manual(values = c("#DE2D26", "#31A354")) +
  labs(x = "Received comprehensive postnatal care", y = "Distance to provider") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r}
#| label: viz-race-insurance-heatmap
#| fig-cap: "Care rates by race/ethnicity and insurance"
#| fig-width: 9
#| fig-height: 5

df %>%
  group_by(race_ethnicity, insurance) %>%
  summarise(care_rate = mean(received_comprehensive_postnatal_care, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = insurance, y = race_ethnicity, fill = care_rate)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(labels = scales::percent_format()) +
  labs(x = "Insurance", y = "Race/Ethnicity", fill = "Care rate") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## 4) Count Outcome and Model Selection

```{r}
#| label: build-count

comorbidity_vars <- c(
  "obesity",
  "multiple_gestation",
  "diabetes",
  "heart_disease",
  "placenta_previa",
  "hypertension",
  "gest_hypertension",
  "preeclampsia"
)

df_model <- df %>%
  mutate(
    comorbidity_count = rowSums(across(all_of(comorbidity_vars)), na.rm = TRUE),
    care_status = factor(received_comprehensive_postnatal_care, levels = c(0, 1), labels = c("No", "Yes"))
  )

df_model %>%
  count(comorbidity_count) %>%
  mutate(prop = n / sum(n))
```

```{r}
#| label: fit-poisson-vs-nb

model_formula <- comorbidity_count ~ care_status + distance_to_provider + insurance + race_ethnicity + age + dependents

poisson_fit <- glm(model_formula, data = df_model, family = poisson(link = "log"))
nb_fit <- MASS::glm.nb(model_formula, data = df_model)

poisson_dispersion <- sum(residuals(poisson_fit, type = "pearson")^2) / poisson_fit$df.residual
model_compare <- tibble(
  model = c("Poisson", "Negative binomial"),
  aic = c(AIC(poisson_fit), AIC(nb_fit))
)

chosen_model <- if (poisson_dispersion > 1.2 && AIC(nb_fit) + 2 < AIC(poisson_fit)) nb_fit else poisson_fit
chosen_model_name <- if (identical(chosen_model, nb_fit)) "Negative binomial" else "Poisson"

list(
  poisson_dispersion_ratio = poisson_dispersion,
  model_aic = model_compare,
  selected_model = chosen_model_name
)
```

```{r}
#| label: model-results

coef_table <- tidy(chosen_model, conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(across(c(estimate, conf.low, conf.high), round, 3)) %>%
  rename(irr = estimate, irr_low = conf.low, irr_high = conf.high)

coef_table %>%
  dplyr::select(term, irr, irr_low, irr_high, p.value) %>%
  arrange(p.value)
```

### Interpretation notes

- Coefficients are reported as incidence rate ratios (IRRs): values above 1 indicate higher expected comorbidity count, while values below 1 indicate lower expected count, conditional on other included covariates.
- Because this is observational synthetic data, these associations should be interpreted as descriptive/associational, not causal.

## 5) Assumptions, Diagnostics, and Limitations

```{r}
#| label: diagnostics

df_model <- df_model %>%
  mutate(
    fitted_mean = fitted(chosen_model),
    pearson_resid = residuals(chosen_model, type = "pearson"),
    deviance_resid = residuals(chosen_model, type = "deviance")
  )

obs_zero_rate <- mean(df_model$comorbidity_count == 0)
pred_zero_rate <- mean(if (chosen_model_name == "Poisson") {
  dpois(0, lambda = df_model$fitted_mean)
} else {
  theta <- chosen_model$theta
  (theta / (theta + df_model$fitted_mean))^theta
})

tibble(
  metric = c("Observed zero rate", "Predicted zero rate"),
  value = c(obs_zero_rate, pred_zero_rate)
)
```

```{r}
#| label: residual-vs-fitted
#| fig-cap: "Pearson residuals vs fitted mean"
#| fig-width: 8
#| fig-height: 4.5

df_model %>%
  ggplot(aes(x = fitted_mean, y = pearson_resid)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "loess", se = FALSE, color = "#045A8D") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#444444") +
  labs(x = "Fitted mean", y = "Pearson residual") +
  theme_minimal(base_size = 12)
```

### Model assumptions and practical limits

- Mean-variance structure: Poisson assumes equality of mean and variance; negative binomial relaxes this with an overdispersion parameter. The model choice above uses both dispersion and AIC.
- Independence: records are assumed conditionally independent; clustering by provider/state could violate this and may require multilevel models.
- Functional form: continuous predictors are modeled linearly on the log scale; nonlinear effects were not explicitly modeled.
- Omitted variables: unmeasured confounding and selection mechanisms can bias associations.
- Outcome definition: `comorbidity_count` is a constructed summary and does not reflect severity weighting across conditions.

## Summary of What Changed

- Loaded and type-normalized the dataset with explicit factor/numeric/binary handling.
- Reported row/column counts, complete missingness profile, descriptive summaries, and data-quality checks.
- Added targeted visualizations connecting postnatal care with distance, insurance, and race/ethnicity (including a race-insurance heatmap).
- Created `comorbidity_count` exactly as specified and fit both Poisson and negative binomial candidates.
- Selected the count model based on dispersion and AIC, then reported exponentiated IRRs with confidence intervals.
- Added diagnostics (zero-rate check, residual-vs-fitted plot) plus non-causal interpretation and limitations.
