---
title: "Tools Comparison (Simple): OpenCode + Gemini 3.1 Pro"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `opencode`
- Model: `gemini-3-1-pro`
- Prompt type: `simple`

## Setup

```{r}
#| label: setup
#| message: false

library(framework)
scaffold()
df <- data_read("data/synthetic/simulated_maternal_health_data.csv")
```

## Prompt

```text
Study this data set (`data/synthetic/simulated_maternal_health_data.csv`), show me descriptive statistics, pick a modeling strategy, and report the model. Be thorough.
```

## Descriptive Statistics

```{r}
#| label: descriptive-stats
#| message: false
#| warning: false

# Overview of variables
glimpse(df)

# Summary of numeric and categorical variables
summary(df)

# Look at the distribution of the outcome variable (Postnatal Care)
table(df$received_comprehensive_postnatal_care, useNA = "always")

# Cross-tabulate with insurance type
table(df$insurance, df$received_comprehensive_postnatal_care)

# Look at age and distance to provider by outcome
df %>%
  group_by(received_comprehensive_postnatal_care) %>%
  summarize(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    mean_distance = mean(distance_to_provider, na.rm = TRUE),
    prop_obesity = mean(obesity, na.rm = TRUE),
    prop_hypertension = mean(hypertension, na.rm = TRUE)
  )
```

## Modeling Strategy

We will use a logistic regression model (General Linear Model with binomial family) to predict the likelihood of receiving comprehensive postnatal care (`received_comprehensive_postnatal_care`) based on demographic and health factors. This allows us to estimate the odds ratios associated with different levels of education, insurance, distance to a provider, and pre-existing health conditions.

```{r}
#| label: modeling
#| message: false

# Convert target variable to binary numeric if it's not already
# It appears to be 0/1 from the glimpse
# Define the logistic regression model
model_care <- glm(
  received_comprehensive_postnatal_care ~ age + edu + insurance + distance_to_provider + obesity + hypertension,
  data = df,
  family = binomial(link = "logit")
)

# Display the summary of the model
summary(model_care)

# Display odds ratios
exp(coef(model_care))
```
