---
title: "Tools Comparison (Detailed): Claude Code + Sonnet 4.6"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false

```

## Run Metadata

- Harness: `claude-code`
- Model: `sonnet-4-6`
- Prompt type: `detailed`

## Setup

```{r}
#| label: setup
#| message: false

library(dplyr)

df <- readr::read_csv(here::here("data/synthetic/simulated_maternal_health_data.csv"))
```

## Prompt

```text
I want you to help me analyze this dataset: `data/synthetic/simulated_maternal_health_data.csv`

My goal is to produce a thorough analysis that is statistically sound and lays out all steps of the analysis.

Please do the following:

1. Load the dataset and normalize columns/types.
   - Use clear snake_case names.
   - Ensure categorical vs numeric types are explicit.

2. Summarize the data.
   - Report row count, column count, missingness, and key descriptive summaries.
   - Flag suspicious values or quality concerns.

3. Create key visualizations.
   - Show relationships for:
     - `received_comprehensive_postnatal_care`
     - `distance_to_provider`
     - `insurance`
     - `race_ethnicity`

4. Build a count outcome and model it.
   - Define `comorbidity_count` as the row-wise sum of:
     - `obesity`,
     - `multiple_gestation`,
     - `diabetes`,
     - `heart_disease`,
     - `placenta_previa`,
     - `hypertension`,
     - `gest_hypertension`,
     - `preeclampsia`.
   - Select an appropriate count-model strategy (Poisson vs negative binomial, with justification).
   - Fit the model and report interpretable results.

5. Report results responsibly.
   - Do not overstate causality.
   - Include model assumptions, diagnostics, and limitations.

Constraints:
- Prefer readable, modular, reproducible code that follows R community idioms.
- Report back a summary of what you changed.
```

## Response

### Step 1: Normalize Column Types

Column names are already in snake_case. We explicitly cast categorical variables
to factors and ensure binary indicators are stored as integers.

```{r}
#| label: normalize-types

# Ordered income levels
income_levels <- c(
  "Under $25,000",
  "$25,000–$49,999",
  "$50,000–$74,999",
  "$75,000–$99,999",
  "$100,000 or more"
)

edu_levels <- c("less_than_hs", "hs", "some_college", "bachelors", "graduate")

comorbidity_cols <- c(
  "obesity", "multiple_gestation", "diabetes", "heart_disease",
  "placenta_previa", "hypertension", "gest_hypertension", "preeclampsia"
)

df <- df |>
  mutate(
    self_report_income = factor(self_report_income, levels = income_levels, ordered = TRUE),
    edu                = factor(edu, levels = edu_levels, ordered = TRUE),
    race_ethnicity     = factor(race_ethnicity),
    insurance          = factor(insurance),
    job_type           = factor(job_type),
    state              = factor(state),
    received_comprehensive_postnatal_care = as.integer(received_comprehensive_postnatal_care),
    across(all_of(comorbidity_cols), as.integer)
  )

glimpse(df)
```

### Step 2: Data Summary and Quality Checks

```{r}
#| label: dimensions

cat("Rows:   ", nrow(df), "\n")
cat("Columns:", ncol(df), "\n")
```

```{r}
#| label: missingness

miss <- colSums(is.na(df))
if (any(miss > 0)) {
  miss[miss > 0]
} else {
  cat("No missing values detected.\n")
}
```

```{r}
#| label: numeric-summaries

df |>
  select(age, dependents, distance_to_provider) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    n      = n(),
    mean   = mean(value, na.rm = TRUE),
    sd     = sd(value, na.rm = TRUE),
    min    = min(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    max    = max(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2, caption = "Numeric variable summaries")
```

```{r}
#| label: categorical-frequencies

cat("--- Insurance ---\n")
print(table(df$insurance, useNA = "ifany"))

cat("\n--- Race/Ethnicity ---\n")
print(table(df$race_ethnicity, useNA = "ifany"))

cat("\n--- Education ---\n")
print(table(df$edu, useNA = "ifany"))

cat("\n--- Received Postnatal Care ---\n")
print(table(df$received_comprehensive_postnatal_care, useNA = "ifany"))
```

```{r}
#| label: quality-flags

# Check distance range (negative values would be suspicious)
cat("Distance to provider — min:", min(df$distance_to_provider, na.rm = TRUE),
    " max:", max(df$distance_to_provider, na.rm = TRUE), "\n")

# Check age range (very young or old would be clinically implausible)
cat("Age — min:", min(df$age, na.rm = TRUE),
    " max:", max(df$age, na.rm = TRUE), "\n")

# Confirm binary columns are 0/1 only
bad_binary <- sapply(comorbidity_cols, function(col) {
  setdiff(unique(df[[col]]), c(0L, 1L))
})
bad_binary[lengths(bad_binary) > 0]
```

### Step 3: Key Visualizations

```{r}
#| label: viz-postnatal-by-insurance
#| fig-cap: "Rate of comprehensive postnatal care by insurance type"

library(ggplot2)

df |>
  group_by(insurance) |>
  summarise(
    pct = mean(received_comprehensive_postnatal_care, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(insurance, pct), y = pct)) +
  geom_col(fill = "#2166ac") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Proportion Receiving Care",
    title = "Comprehensive Postnatal Care by Insurance Type"
  )
```

```{r}
#| label: viz-postnatal-by-race
#| fig-cap: "Rate of comprehensive postnatal care by race/ethnicity"

df |>
  group_by(race_ethnicity) |>
  summarise(
    pct = mean(received_comprehensive_postnatal_care, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(race_ethnicity, pct), y = pct)) +
  geom_col(fill = "#d6604d") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Proportion Receiving Care",
    title = "Comprehensive Postnatal Care by Race/Ethnicity"
  )
```

```{r}
#| label: viz-distance-distribution
#| fig-cap: "Distribution of distance to provider by postnatal care receipt"

df |>
  mutate(
    care_status = ifelse(
      received_comprehensive_postnatal_care == 1,
      "Received Care", "Did Not Receive"
    )
  ) |>
  ggplot(aes(x = distance_to_provider, fill = care_status)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("Received Care" = "#4dac26", "Did Not Receive" = "#b8860b")) +
  labs(
    x = "Distance to Provider (miles)",
    y = "Density",
    fill = NULL,
    title = "Distance to Provider by Postnatal Care Status"
  )
```

```{r}
#| label: viz-distance-by-race
#| fig-cap: "Distance to provider by race/ethnicity"

df |>
  ggplot(aes(x = reorder(race_ethnicity, distance_to_provider, median),
             y = distance_to_provider)) +
  geom_boxplot(fill = "#74add1", outlier.alpha = 0.1) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Distance to Provider (miles)",
    title = "Distance to Provider by Race/Ethnicity"
  )
```

### Step 4: Comorbidity Count Outcome

#### 4a. Create the Outcome Variable

```{r}
#| label: create-comorbidity-count

df <- df |>
  mutate(comorbidity_count = rowSums(across(all_of(comorbidity_cols))))

cat("Comorbidity count distribution:\n")
print(table(df$comorbidity_count))
cat("\nMean:", round(mean(df$comorbidity_count), 3),
    "  Variance:", round(var(df$comorbidity_count), 3), "\n")
```

```{r}
#| label: viz-comorbidity
#| fig-cap: "Distribution of comorbidity count"

ggplot(df, aes(x = comorbidity_count)) +
  geom_bar(fill = "#5e3c99") +
  labs(
    x = "Number of Comorbidities",
    y = "Count",
    title = "Distribution of Comorbidity Count"
  )
```

#### 4b. Model Selection: Poisson vs Negative Binomial

For count outcomes we first consider Poisson regression. Poisson assumes
mean = variance; if the variance substantially exceeds the mean (overdispersion),
a negative binomial model—which estimates an extra dispersion parameter θ—is
preferred.

```{r}
#| label: overdispersion-check

dispersion_ratio <- var(df$comorbidity_count) / mean(df$comorbidity_count)
cat("Mean:             ", round(mean(df$comorbidity_count), 3), "\n")
cat("Variance:         ", round(var(df$comorbidity_count), 3), "\n")
cat("Variance/Mean:    ", round(dispersion_ratio, 3), "\n")
cat("Overdispersed?    ", ifelse(dispersion_ratio > 1.5, "Yes — favour NB", "Marginal / No"), "\n")
```

```{r}
#| label: fit-poisson

fit_pois <- glm(
  comorbidity_count ~ age + edu + race_ethnicity + insurance +
    distance_to_provider + self_report_income + dependents,
  data   = df,
  family = poisson(link = "log")
)

# Residual deviance / df as a quick overdispersion diagnostic
cat("Poisson residual deviance / df:",
    round(fit_pois$deviance / fit_pois$df.residual, 3), "\n")
```

```{r}
#| label: fit-negbin

library(MASS)

fit_nb <- glm.nb(
  comorbidity_count ~ age + edu + race_ethnicity + insurance +
    distance_to_provider + self_report_income + dependents,
  data = df
)

cat("Poisson AIC:           ", round(AIC(fit_pois), 1), "\n")
cat("Negative Binomial AIC: ", round(AIC(fit_nb), 1), "\n")
```

```{r}
#| label: select-model

use_nb <- AIC(fit_nb) < AIC(fit_pois)
best_fit   <- if (use_nb) fit_nb   else fit_pois
model_name <- if (use_nb) "Negative Binomial" else "Poisson"
cat("Selected model:", model_name, "\n")
```

#### 4c. Interpretable Results

Coefficients are exponentiated to Incidence Rate Ratios (IRRs). An IRR > 1
indicates a higher expected comorbidity count; IRR < 1 indicates a lower count,
holding all other predictors constant.

```{r}
#| label: irr-table

irr    <- exp(coef(best_fit))
irr_ci <- exp(confint.default(best_fit))

data.frame(
  IRR      = irr,
  CI_lower = irr_ci[, 1],
  CI_upper = irr_ci[, 2]
) |>
  knitr::kable(digits = 3, caption = paste("IRRs from", model_name, "Model"))
```

#### 4d. Model Diagnostics

```{r}
#| label: diagnostics
#| fig-cap: "Residual diagnostic plots for the selected model"

par(mfrow = c(2, 2))
plot(best_fit)
par(mfrow = c(1, 1))
```

### Step 5: Responsible Reporting

#### Summary of Changes

- Converted `self_report_income` and `edu` to ordered factors with substantively
  meaningful level sequences.
- Cast `race_ethnicity`, `insurance`, `job_type`, and `state` to unordered factors.
- Ensured all eight binary comorbidity indicators are stored as integers.
- Created `comorbidity_count` as a row-wise sum of those eight indicators.
- Chose between Poisson and negative binomial via variance/mean ratio and AIC
  comparison; reported exponentiated coefficients (IRRs) with 95% confidence
  intervals.

#### Limitations and Cautions

1. **Association, not causation.** This is a cross-sectional analysis. The IRRs
   describe statistical associations conditional on included covariates, not
   causal effects.
2. **Unmeasured confounding.** Variables such as access to preventive care,
   health literacy, and provider quality are absent from the model and may
   confound observed associations.
3. **Clustering not modelled.** Patients share `provider_id` and `state`, which
   likely induces within-cluster correlation. Mixed models or clustered standard
   errors would better account for this structure.
4. **Self-reported income.** Broad income brackets introduce measurement
   imprecision and are subject to recall and social-desirability bias.
5. **Synthetic data.** The dataset was simulated. Results illustrate the analytic
   workflow and should not be interpreted as real-world epidemiological evidence.
6. **Model specification.** Predictors were selected from the prompt rather than
   from a domain-informed causal model (e.g., a directed acyclic graph). Some
   covariates may be mediators of the exposure–outcome relationship.
